---
title: "Niche Composition: Pseudobulk Treatment Comparison"
subtitle: "Per-brain aggregation of plaque-proximal cellular niches, Adu vs IgG"
date: last-modified
---

# Setup

```{r}
#| label: setup
source("R/setup.R")
library(patchwork)

data <- load_slides_and_distances("full")
slide1 <- data$slide1; slide2 <- data$slide2
cell_distances_1 <- data$distances_1; cell_distances_2 <- data$distances_2
cell_distances_all <- data$distances_all

sample_lookup <- cell_distances_all |> dplyr::select(cell, sample_ID, Treatment.Group)
```

# Motivation

The treatment comparison in notebook 11 (`11_neighborhood_enrichment.qmd`) pools all cells from
Adu-treated brains and compares them against all cells from IgG-treated brains, treating each cell
as an independent observation. This inflates effective sample size and ignores the nested structure
of the experiment: 3 biological replicates per group.

Here we take a **pseudobulk approach**: niche proportions are first averaged within each brain, then
compared across the 3 Adu brains vs 3 IgG brains. This makes the biological replication explicit
and is the appropriate unit of inference.

> **Note on statistical power.** With n = 3 replicates per group, all tests are severely
> underpowered. Welch's t-test is used here because, unlike the Wilcoxon rank-sum test (minimum
> two-sided p-value of 0.1 at n = 3 vs n = 3), it can in principle reach p < 0.05 when
> between-group differences are large relative to within-group variance. Normality cannot be
> verified at n = 3, so results should be interpreted as exploratory and directional.

# Compute per-brain niche composition

For every plaque-proximal cell (< 50 µm from the nearest plaque), the 15 nearest spatial
neighbours are identified within the same tissue slide, and the proportion of each annotated cell
type among those neighbours is recorded. Proportions are then **averaged to the brain level**,
yielding 6 brain-level observations per (plaque type × neighbour cell type) combination.

```{r}
#| label: compute-niche
compute_niche_composition <- function(seurat_obj, cell_distances, threshold = 50) {
  proximal_info <- cell_distances |>
    mutate(
      nearest_type = if_else(dist_caa < dist_paren, "CAA", "Parenchymal"),
      min_dist     = pmin(dist_caa, dist_paren)
    ) |>
    filter(min_dist < threshold)

  if (nrow(proximal_info) == 0) return(tibble())

  coords <- GetTissueCoordinates(seurat_obj) |> as.data.frame()
  meta   <- seurat_obj@meta.data |> as_tibble(rownames = "cell")

  all_mat  <- coords |> dplyr::select(x, y) |> as.matrix()
  prox_idx <- match(proximal_info$cell, coords$cell)
  prox_idx <- prox_idx[!is.na(prox_idx)]

  nn_res <- nn2(data = all_mat, query = all_mat[prox_idx, ], k = 16)

  map(seq_along(prox_idx), \(i) {
    neighbour_idx <- nn_res$nn.idx[i, -1]
    tibble(
      focal_cell     = coords$cell[prox_idx[i]],
      neighbour_type = meta$annotatedclusters[neighbour_idx]
    )
  }) |>
    bind_rows() |>
    count(focal_cell, neighbour_type) |>
    complete(focal_cell, neighbour_type, fill = list(n = 0L)) |>
    group_by(focal_cell) |>
    mutate(prop = n / sum(n)) |>
    ungroup() |>
    left_join(
      proximal_info |> dplyr::select(cell, nearest_type),
      by = c("focal_cell" = "cell")
    ) |>
    filter(!is.na(nearest_type))
}

niche_1 <- compute_niche_composition(slide1, cell_distances_1)
niche_2 <- compute_niche_composition(slide2, cell_distances_2)

niche_all <- bind_rows(
  niche_1 |> mutate(slide = 1L),
  niche_2 |> mutate(slide = 2L)
) |>
  left_join(sample_lookup, by = c("focal_cell" = "cell"))
```

## Pseudobulk aggregation

```{r}
#| label: pseudobulk-aggregation
niche_brain <- niche_all |>
  group_by(sample_ID, Treatment.Group, nearest_type, neighbour_type) |>
  summarise(mean_prop = mean(prop, na.rm = TRUE), .groups = "drop")
```

# Treatment comparison

## Dot plots: per-brain niche proportions

Each point represents one brain (n = 3 per group). Crossbars show group means. Plotting individual
brains makes consistency (or inconsistency) of treatment effects across replicates directly visible.

```{r}
#| label: dotplot-treatment
#| column: page
#| fig-height: 8
#| fig-width: 14

niche_brain |>
  ggplot(aes(x = neighbour_type, y = mean_prop, colour = Treatment.Group)) +
  geom_point(
    position = position_dodge(width = 0.6),
    size = 2.5, alpha = 0.85
  ) +
  stat_summary(
    aes(group = Treatment.Group),
    fun = mean, geom = "crossbar",
    width = 0.45,
    position = position_dodge(width = 0.6),
    linewidth = 0.5
  ) +
  scale_colour_manual(values = c(Adu = "#E05A4E", IgG = "#4E79A7")) +
  facet_wrap(~ nearest_type, ncol = 1) +
  labs(
    title = "Niche composition: Adu vs IgG (pseudobulk, n = 3 per group)",
    x = "Neighbour cell type",
    y = "Mean niche proportion",
    colour = "Treatment"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Welch's t-test (exploratory)

```{r}
#| label: ttest-results
ttest_results <- niche_brain |>
  group_by(nearest_type, neighbour_type) |>
  summarise(
    mean_Adu = mean(mean_prop[Treatment.Group == "Adu"], na.rm = TRUE),
    mean_IgG = mean(mean_prop[Treatment.Group == "IgG"], na.rm = TRUE),
    delta    = mean_Adu - mean_IgG,
    log2fc   = log2((mean_Adu + 1e-6) / (mean_IgG + 1e-6)),
    p_value  = tryCatch(
      t.test(
        mean_prop[Treatment.Group == "Adu"],
        mean_prop[Treatment.Group == "IgG"]
      )$p.value,
      error = \(e) NA_real_
    ),
    .groups = "drop"
  ) |>
  mutate(p_adj = p.adjust(p_value, method = "BH")) |>
  arrange(nearest_type, p_value)

ttest_results |>
  mutate(across(c(mean_Adu, mean_IgG, delta, log2fc), \(x) round(x, 4))) |>
  mutate(across(c(p_value, p_adj), \(x) round(x, 3))) |>
  kableExtra::kbl(
    caption = "Pseudobulk Welch's t-test: Adu vs IgG niche proportions (n = 3 per group)",
    col.names = c(
      "Plaque type", "Neighbour type",
      "Mean Adu", "Mean IgG", "Δ (Adu − IgG)", "log₂FC", "p-value", "p adj (BH)"
    )
  ) |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )
```

## Difference heatmap (Adu − IgG)

Red tiles indicate cell types proportionally more abundant in Adu niche; blue tiles indicate
depletion relative to IgG. The colour scale shows the raw difference in mean niche proportion.

```{r}
#| label: delta-heatmap
#| column: page
#| fig-height: 3.5
#| fig-width: 12

ttest_results |>
  ggplot(aes(x = neighbour_type, y = nearest_type, fill = delta)) +
  geom_tile(colour = "white") +
  scale_fill_gradient2(
    low = "#4E79A7", mid = "white", high = "#E05A4E", midpoint = 0,
    name = "Δ proportion\n(Adu − IgG)"
  ) +
  labs(
    title = "Treatment difference in niche composition (pseudobulk)",
    subtitle = "Δ = mean(Adu) − mean(IgG) per brain; n = 3 per group",
    x = "Neighbour cell type",
    y = "Nearest plaque type"
  ) +
  theme_minimal(base_size = 15) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
