---
title: "Bivariate Spatial Cross-Correlation"
subtitle: "Lee's L per sample (brain) for treatment comparison — kNN(15) spatial weights"
date: last-modified
---

# Setup

```{r}
#| label: setup
source("R/setup.R")
library(spdep)

data   <- load_slides_and_distances("full")
slide1 <- data$slide1
slide2 <- data$slide2

## Identify cell-types of interest

ct_astro <- "Astrocyte"
ct_oligo <- "Oligodendrocyte"
ct_micro <- "Microglia"


```

# Hypothesis

The three-layer glial response model proposes a spatial relay: microglia activation
(Layer 1) drives astrocyte amplification (Layer 2), which in turn triggers
oligodendrocyte remodeling (Layer 3).

If this relay is spatially structured rather than three independent regional responses,
then activated microglia and reactive astrocytes should be spatially co-located: where
one finds a microglial *Ccl12* hotspot, one should also find a *Cxcl10*-expressing
astrocyte hotspot nearby.

**Lee's L statistic** is the bivariate generalisation of Moran's I: it quantifies whether
two spatial variables are jointly clustered. We use it to test the following pairs:

| Pair | Gene X | Cell type X | Gene Y | Cell type Y | Hypothesis |
|------|--------|------------|--------|------------|------------|
| 1 | *Ccl12* | Microglia | *Cxcl10* | Astrocytes | Layer 1 → Layer 2 co-activation |
| 2 | *Cxcl10* | Astrocytes | *Opcml* | Oligodendrocytes | Layer 2 → Layer 3 relay |
| 3 | *Ccl12* | Microglia | *Opcml* | Oligodendrocytes | Long-range Layer 1 → Layer 3 |

For each pair we create two per-cell expression vectors: the gene's raw count if the
cell belongs to the relevant cell type, zero otherwise. Lee's L of these two vectors
tests whether the spatial patterns are concordant (L > 0) or discordant (L < 0) beyond
what is expected by chance. Significance is assessed by Monte Carlo permutation (n = 499
shuffles) while keeping cell coordinates fixed.

## Gene selection rationale

Each gene was chosen as the strongest spatially-validated marker for its layer in this
dataset, based on the upstream plaque-proximity and pseudobulk DE analyses:

- ***Ccl12* (Microglia — Layer 1)**: the most statistically significant gene enriched in
  the proximity of parenchymal plaques in Aducanumab-treated brains (notebook 8,
  pseudobulk DESeq2 in plaque-proximal microglia). It also appears as the most
  cross-regionally consistent microglial signal in the DEG_consolidated regional model,
  significant in both hippocampus and white matter microglia.

- ***Cxcl10* (Astrocytes — Layer 2)**: a marker of astrocyte subcluster 6, which is
  expanded in Aducanumab-treated brains relative to IgG controls. *Cxcl10* encodes an
  interferon-γ-inducible chemokine (IP-10) produced by reactive astrocytes, linking
  microglial inflammatory signalling to an astrocyte activation state that is
  treatment-sensitive. This makes it the most informative astrocyte gene to test for
  spatial co-localization with microglial *Ccl12*.

- ***Opcml* (Oligodendrocytes — Layer 3)**: the most statistically significant gene
  enriched near parenchymal plaques upon Aducanumab treatment in the oligodendrocyte
  gene-distance correlation analysis (notebook 12). It is an established oligodendrocyte
  lineage marker and its plaque-proximal upregulation provides the strongest Layer 3
  spatial signal in this dataset.

# Functions



```{r}
#| label: build-cell-type-expression-vectors
# Build a per-cell expression vector for gene G in cell type T:
# value = raw count if cell is of type T, 0 otherwise.
# cells_ordered must match the order of nodes in the spatial weight matrix (lw).
build_ct_expr_vector <- function(seurat_obj, gene, cell_types, cells_ordered = colnames(seurat_obj)) {
  meta       <- seurat_obj@meta.data
  counts_mat <- LayerData(seurat_obj, assay = DefaultAssay(seurat_obj), layer = "counts")

  n    <- length(cells_ordered)
  expr <- numeric(n)

  if (gene %in% rownames(counts_mat)) {
    in_mat       <- cells_ordered %in% colnames(counts_mat)
    expr[in_mat] <- as.numeric(counts_mat[gene, cells_ordered[in_mat]])
  }

  ct_cells <- rownames(meta)[meta$annotatedclusters %in% cell_types]
  expr[!cells_ordered %in% ct_cells] <- 0
  expr
}
```

```{r}
#| label: define-bivariate-lee
# Run Lee's L with Monte Carlo permutation for one gene pair.
run_lee <- function(seurat_obj, lw, gene_x, ct_x, gene_y, ct_y, nsim = 499) {
  lw_cells <- attr(lw$neighbours, "region.id")
  x <- build_ct_expr_vector(seurat_obj, gene_x, ct_x, lw_cells)
  y <- build_ct_expr_vector(seurat_obj, gene_y, ct_y, lw_cells)

  if (sum(x > 0) < 10 || sum(y > 0) < 10) {
    return(tibble(
      gene_x = gene_x, ct_x = paste(ct_x, collapse = "/"),
      gene_y = gene_y, ct_y = paste(ct_y, collapse = "/"),
      lee_L  = NA_real_, pval = NA_real_,
      n_x    = sum(x > 0), n_y = sum(y > 0)
    ))
  }

  mc_res <- lee.mc(x, y, listw = lw, nsim = nsim, zero.policy = TRUE)
  tibble(
    gene_x = gene_x, ct_x = paste(ct_x, collapse = "/"),
    gene_y = gene_y, ct_y = paste(ct_y, collapse = "/"),
    lee_L  = mc_res$statistic,
    pval   = mc_res$p.value,
    n_x    = sum(x > 0),
    n_y    = sum(y > 0)
  )
}
```

# Run Lee's L per sample

Each of the 6 brains is processed independently: subset the slide-level Seurat object
to one sample's cells, rebuild a kNN(15) spatial weight matrix from that sample's cell
coordinates, then run Lee's L for the 3 relay pairs. This gives 18 rows (6 samples x 3
pairs) and enables direct Adu vs IgG comparison (n = 3 per group).

```{r}
#| label: run-bivariate-lee
#| eval: false
relay_pairs <- list(
  list(gene_x = "Ccl12", ct_x = ct_micro, gene_y = "Cxcl10", ct_y = ct_astro),
  list(gene_x = "Cxcl10", ct_x = ct_astro, gene_y = "Opcml", ct_y = ct_oligo),
  list(gene_x = "Ccl12", ct_x = ct_micro, gene_y = "Opcml", ct_y = ct_oligo)
)

t0 <- Sys.time()

lee_results <- map(seq_len(nrow(sample_dict)), \(i) {
  t_start <- Sys.time()
  sid       <- sample_dict$sample[i]
  treatment <- sample_dict$treatment[i]
  slide_num <- sample_dict$slide[i]
  slide_obj <- if (slide_num == 1) slide1 else slide2

  # Subset to this sample's cells
  cells_i <- colnames(slide_obj)[slide_obj$sample_ID == sid]
  sub_obj <- subset(slide_obj, cells = cells_i)

  # Build per-sample kNN(15) spatial weight matrix
  coords <- GetTissueCoordinates(sub_obj) |>
    as.data.frame() |>
    dplyr::select(x, y) |>
    as.matrix()
  nn      <- nn2(coords, coords, k = 16)
  nb_list <- map(seq_len(nrow(coords)), \(j) as.integer(nn$nn.idx[j, -1]))
  class(nb_list) <- "nb"
  attr(nb_list, "region.id") <- cells_i
  lw_i <- nb2listw(nb_list, style = "W", zero.policy = TRUE)

  res <- map(relay_pairs, \(pair) {
    run_lee(sub_obj, lw_i, pair$gene_x, pair$ct_x, pair$gene_y, pair$ct_y) |>
      mutate(sample_ID = sid, treatment = treatment)
  }) |>
    bind_rows()

  elapsed <- as.numeric(difftime(Sys.time(), t_start, units = "mins"))
  message(sprintf("[%d/%d] %s (%s) — %.1f min",
                  i, nrow(sample_dict), sid, treatment, elapsed))
  res
}) |>
  bind_rows()

message(sprintf("Done. Total: %.1f min", as.numeric(difftime(Sys.time(), t0, units = "mins"))))

# write_csv(lee_results, "spatial/lee_bivariate.csv")
```

# Results

## Table

`n_x` and `n_y` are the number of cells with non-zero raw counts for gene X and gene Y,
respectively, **within their designated cell type**. Because the expression vectors are
zeroed out for all cells outside the target cell type, these counts reflect only the
spatially active signal entering Lee's L — they are the effective sample sizes for each
variable. A very small n (< 10) causes the test to be skipped (reported as NA).

```{r}
#| label: lee-results-table
lee_results <- read_csv("spatial/lee_bivariate.csv")

lee_results |>
  mutate(
    pair  = paste0(gene_x, " (", ct_x, ")\n× ", gene_y, " (", ct_y, ")"),
    lee_L = round(lee_L, 4),
    pval  = formatC(pval, format = "f", digits = 3)
  ) |>
  select(pair, sample_ID, treatment, lee_L, pval, n_x, n_y) |>
  kbl(
    caption = "Bivariate Lee's L — three-layer relay pairs (per sample)",
    col.names = c("Pair", "Sample", "Treatment", "Lee's L", "p-value (MC)",
                  "n_x (expressing cells)", "n_y (expressing cells)")
  ) |>
  kable_styling("striped", full_width = FALSE)
```

## Dot plot

```{r}
#| label: lee-results-dotplot
#| column: page
#| fig-height: 5
#| fig-width: 10
#| fig-cap: "Lee's L per brain, faceted by relay pair. Each point is one brain; colour indicates treatment group. Positive L supports spatial co-activation; dashed line at L = 0 (null). Adu brains above IgG would indicate treatment-enhanced relay coupling."

lee_plot <- lee_results |>
  mutate(
    pair = paste0(gene_x, " (", ct_x, ")\n× ", gene_y, " (", ct_y, ")"),
    treatment = factor(treatment, levels = c("IgG", "Adu"))
  )

ggplot(lee_plot, aes(x = treatment, y = lee_L, colour = treatment, size = -log10(pval))) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  geom_jitter(width = 0.1, height = 0) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.4, linewidth = 0.5,
               colour = "black") +
  facet_wrap(~ pair, scales = "free_y") +
  scale_colour_manual(values = c("IgG" = "#4A90D9", "Adu" = "#E07B39")) +
  scale_size_continuous(name = "-log10(p)", range = c(1, 6)) +
  labs(
    title  = "Bivariate Lee's L ",
    x      = NULL,
    y      = "Lee's L",
    colour = "Treatment"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    strip.text = element_text(size = 12),
    panel.grid.minor = element_blank(),
    #panel.grid.major = element_blank(),
    legend.position = "bottom"
  )
```

# Interpretation

Running Lee's L per sample (brain) rather than per slide enables treatment-level
comparison with n = 3 biological replicates per group.

A **positive and significant Lee's L** for the *Ccl12*(microglia) × *Cxcl10*(astrocyte)
pair indicates that regions of microglial *Ccl12* activation are spatially concordant
with *Cxcl10*-expressing astrocytes — direct spatial evidence for the Layer 1 → Layer 2
relay. A positive L for *Cxcl10*(astrocyte) × *Opcml*(oligodendrocyte) supports the
Layer 2 → Layer 3 step. If both pairs show L > 0 but the *Ccl12* × *Opcml* pair
(skipping Layer 2) shows weaker L, this provides spatial evidence for a sequential
relay rather than independent parallel responses.

If Aducanumab-treated brains show consistently higher Lee's L than IgG controls, this
would indicate that anti-amyloid treatment strengthens the spatial coupling between
glial response layers — the relay is not merely present but treatment-modulated.
