---
title: "Gene-distance correlation — astrocytes and oligodendrocytes"
date: last-modified
---


Here we extend the same gene-distance correlation approach to **astrocytes** and
**oligodendrocytes** to test whether Layers 2 and 3 also show plaque-proximity
gradients:

- **Layer 2 — Amplification (Astrocytes)**: Ifitm3, Ccl2, Ccl12, Cxcl10, Tnf, Agt
- **Layer 3 — Remodeling (Oligodendrocytes)**: Abca1, Abcg1, Apod, Srebf1, Got1↓, Ldhb↓



# Setup and data loading

```{r}
#| label: setup
source("R/setup.R")
library(ggrepel)
library(patchwork)
library(ggVennDiagram)

data <- load_slides_and_distances("full")
slide1 <- data$slide1; slide2 <- data$slide2
cell_distances_all <- data$distances_all


## Identify cell-types of interest

ct_astro <- "Astrocyte"
ct_oligo <- "Oligodendrocyte"
```



## Subset objects and distances

```{r}
#| label: subset-objects
# Subset Seurat objects to target cell type
subset_by_type <- function(slide_list, cell_types) {
  map(slide_list, \(obj) {
    keep <- obj$annotatedclusters %in% cell_types
    subset(obj, cells = colnames(obj)[keep])
  })
}

slides_astro <- subset_by_type(list(slide1, slide2), ct_astro)
slides_oligo <- subset_by_type(list(slide1, slide2), ct_oligo)

# Subset distance table to matching cells
dist_astro <- cell_distances_all |>
  filter(cell %in% unlist(map(slides_astro, colnames)))

dist_oligo <- cell_distances_all |>
  filter(cell %in% unlist(map(slides_oligo, colnames)))

message(sprintf("Astrocytes: %d cells", nrow(dist_astro)))
message(sprintf("Oligodendrocytes: %d cells", nrow(dist_oligo)))
```

## Cell counts per group

```{r}
#| label: cell-counts
build_count_table <- function(dist_df, label) {
  dist_df |>
    mutate(
      prox_caa   = if_else(dist_caa < 50,   "proximal", "distal"),
      prox_paren = if_else(dist_paren < 50,  "proximal", "distal")
    ) |>
    group_by(sample_ID, Treatment.Group) |>
    summarise(
      n_total       = n(),
      n_prox_caa    = sum(prox_caa == "proximal"),
      n_prox_paren  = sum(prox_paren == "proximal"),
      .groups = "drop"
    ) |>
    mutate(cell_type = label)
}

bind_rows(
  build_count_table(dist_astro, "Astrocytes"),
  build_count_table(dist_oligo, "Oligodendrocytes")
) |>
  kbl(caption = "Cell counts per sample and proximity group") |>
  kable_styling("striped", full_width = FALSE)
```

# Gene-distance correlation functions

`compute_gene_distance_correlation()` is centralised in `R/setup.R` (sourced in setup chunk). Cell-type-specific analyses use `min_expressing_cells = 50` (lower than the default of 100 to accommodate smaller cell populations).

```{r}
#| label: define-plot-treatment-scatter
plot_treatment_scatter <- function(cor_adu, cor_igg, title) {
  df <- inner_join(
    cor_adu |> select(gene, rho_flip_adu = rho_flip, sig_adu = sig),
    cor_igg |> select(gene, rho_flip_igg = rho_flip, sig_igg = sig),
    by = "gene"
  ) |>
    mutate(
      concordance = case_when(
        sig_adu == "enriched_near" & sig_igg == "enriched_near" ~ "both near",
        sig_adu == "enriched_far"  & sig_igg == "enriched_far"  ~ "both far",
        sig_adu != "ns"            & sig_igg != "ns"            ~ "discordant",
        sig_adu != "ns"            | sig_igg != "ns"            ~ "one sig",
        TRUE                                                     ~ "ns"
      )
    )

  colour_map <- c(
    "both near"  = "#E07B39",
    "both far"   = "#7B68AE",
    "discordant" = "#C0392B",
    "one sig"    = "#5DADE2",
    "ns"         = "grey80"
  )

  label_df <- df |> filter(concordance %in% c("both near", "both far", "discordant"))

  ggplot(df, aes(x = rho_flip_igg, y = rho_flip_adu)) +
    geom_point(aes(colour = concordance), size = 1.8, alpha = 0.6) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
    geom_hline(yintercept = 0, colour = "grey70", linewidth = 0.3) +
    geom_vline(xintercept = 0, colour = "grey70", linewidth = 0.3) +
    geom_text_repel(
      data = label_df, aes(label = gene, colour = concordance),
      size = 3.5, fontface = "italic", max.overlaps = 20, show.legend = FALSE
    ) +
    scale_colour_manual(values = colour_map) +
    labs(title = title, x = "rho_flip — IgG", y = "rho_flip — Aducanumab", colour = NULL) +
    theme_minimal(base_size = 13) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank(),
          panel.grid.major = element_line(colour = "grey92"))
}
```

```{r}
#| label: define-plot-correlation
plot_gene_distance_correlation <- function(
    results,
    title          = "Gene-distance correlation",
    n_top          = 5,
    genes_to_label = NULL
) {
  top_pos    <- results |> filter(sig == "enriched_near") |> slice_min(rank, n = n_top)
  top_neg    <- results |> filter(sig == "enriched_far")  |> slice_max(rank, n = n_top)
  auto_label <- bind_rows(top_pos, top_neg)

  if (!is.null(genes_to_label)) {
    user_genes <- results |> filter(gene %in% genes_to_label)
    auto_label <- bind_rows(auto_label, user_genes) |> distinct(gene, .keep_all = TRUE)
  }

  colour_map <- c(
    "enriched_near" = "#E07B39",
    "enriched_far"  = "#7B68AE",
    "ns"            = "grey70"
  )

  ggplot(results, aes(x = rank, y = rho_flip)) +
    geom_point(aes(colour = sig), size = 2, alpha = 0.5) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
    geom_text_repel(
      data        = auto_label,
      aes(label = gene, colour = sig),
      size        = 5,
      fontface    = "italic",
      box.padding = 0.4,
      max.overlaps = 20,
      show.legend = FALSE,
      seed = 23
    ) +
    scale_colour_manual(
      values = colour_map,
      labels = c(
        "enriched_near" = "Enriched near plaque (padj < 0.05)",
        "enriched_far"  = "Enriched far from plaque (padj < 0.05)",
        "ns"            = "Not significant"
      )
    ) +
    labs(
      title  = title,
      x      = "Gene rank",
      y      = "Spearman rho (flipped)",
      colour = NULL
    ) +
    theme_minimal(base_size = 14) +
    theme(
      legend.position    = "bottom",
      panel.grid.minor   = element_blank(),
      panel.grid.major.x = element_blank()
    )
}
```

# Astrocyte gene-distance correlations

Candidate genes from DEG_consolidated model (Layer 2 — Amplification):
- Near plaque: Ifitm3, Ccl2, Ccl12, Cxcl10, Agt, Gfap, Vim, S100b
- Far from plaque: homeostatic astrocyte markers

```{r}
#| label: run-astro-correlations
#| eval: false
astro_candidate_genes <- c(
  "Ifitm3", "Ccl2", "Ccl12", "Cxcl10", "Agt",
  "Gfap", "Vim", "S100b", "Apod", "Abca1", "Plin4"
)

cor_caa_astro   <- compute_gene_distance_correlation(
  slides_astro, dist_astro, distance_col = "dist_caa",   min_expressing_cells = 50)
cor_paren_astro <- compute_gene_distance_correlation(
  slides_astro, dist_astro, distance_col = "dist_paren", min_expressing_cells = 50)
cor_any_astro   <- compute_gene_distance_correlation(
  slides_astro, dist_astro, distance_col = "dist_any",   min_expressing_cells = 50)

# qs2::qs_save(cor_caa_astro,   "spatial/astro_cor_caa.qs2")
# qs2::qs_save(cor_paren_astro, "spatial/astro_cor_paren.qs2")
# qs2::qs_save(cor_any_astro,   "spatial/astro_cor_any.qs2")
```

```{r}
#| label: run-astro-correlations-by-treatment
#| eval: false
dist_astro_adu <- dist_astro |> filter(Treatment.Group == "Adu")
dist_astro_igg <- dist_astro |> filter(Treatment.Group == "IgG")

for (trt in c("adu", "igg")) {
  dist_trt <- if (trt == "adu") dist_astro_adu else dist_astro_igg
  for (dcol in c("dist_caa", "dist_paren", "dist_any")) {
    res    <- compute_gene_distance_correlation(slides_astro, dist_trt,
                distance_col = dcol, min_expressing_cells = 50)
    suffix <- sub("dist_", "", dcol)
    qs2::qs_save(res, paste0("spatial/astro_cor_", suffix, "_", trt, ".qs2"))
  }
}
```

```{r}
#| label: load-astro-correlations
astro_candidate_genes <- c(
  "Ifitm3", "Ccl2", "Ccl12", "Cxcl10", "Agt",
  "Gfap", "Vim", "S100b", "Apod", "Abca1", "Plin4"
)

cor_caa_astro   <- qs2::qs_read("spatial/astro_cor_caa.qs2")
cor_paren_astro <- qs2::qs_read("spatial/astro_cor_paren.qs2")
cor_any_astro   <- qs2::qs_read("spatial/astro_cor_any.qs2")

astro_cor_caa_adu   <- qs2::qs_read("spatial/astro_cor_caa_adu.qs2")
astro_cor_caa_igg   <- qs2::qs_read("spatial/astro_cor_caa_igg.qs2")
astro_cor_paren_adu <- qs2::qs_read("spatial/astro_cor_paren_adu.qs2")
astro_cor_paren_igg <- qs2::qs_read("spatial/astro_cor_paren_igg.qs2")
astro_cor_any_adu   <- qs2::qs_read("spatial/astro_cor_any_adu.qs2")
astro_cor_any_igg   <- qs2::qs_read("spatial/astro_cor_any_igg.qs2")
```

## Rank plots — astrocytes

```{r}
#| label: plot-astro-caa-paren
#| column: page
#| fig-height: 5
#| fig-width: 10
#| fig-cap: "Spearman gene-distance correlation for astrocytes. Positive rho_flip = enriched near plaque. Top 5 enriched-near and enriched-far genes are automatically labeled; candidate Layer 2 genes (Ifitm3, Ccl2, Ccl12, Cxcl10, Agt, Gfap, Vim) are additionally labeled if significant."

p_caa_astro   <- plot_gene_distance_correlation(
  cor_caa_astro,
  title          = "Astrocytes — CAA",
  genes_to_label = astro_candidate_genes
)

p_paren_astro <- plot_gene_distance_correlation(
  cor_paren_astro,
  title          = "Astrocytes — Parenchymal",
  genes_to_label = astro_candidate_genes
)

(p_caa_astro | p_paren_astro) +
  plot_annotation(title = "Astrocyte gene-distance correlation") +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

## Top astrocyte genes

```{r}
#| label: astro-top-tables
show_top <- function(results, label, n = 10) {
  bind_rows(
    results |> filter(sig == "enriched_near") |> slice_min(rank, n = n),
    results |> filter(sig == "enriched_far")  |> slice_max(rank, n = n)
  ) |>
    select(gene, rho, rho_flip, pval, padj, rank, sig) |>
    mutate(
      across(c(rho, rho_flip), \(x) round(x, 4)),
      across(c(pval, padj), \(x) ifelse(x < 1e-300, "< 1e-300", formatC(x, format = "e", digits = 2)))
    ) |>
    kbl(caption = label) |>
    kable_styling("striped", full_width = FALSE)
}

show_top(cor_caa_astro,   "Top genes — astrocytes, distance to CAA")
show_top(cor_paren_astro, "Top genes — astrocytes, distance to parenchymal")
```

## Treatment-split rank plots — astrocytes

```{r}
#| label: plot-astro-caa-treatment
#| column: page
#| fig-height: 5
#| fig-width: 10
plot_gene_distance_correlation(astro_cor_caa_adu,
  title = "Astrocytes — CAA, Aducanumab", genes_to_label = astro_candidate_genes) |
plot_gene_distance_correlation(astro_cor_caa_igg,
  title = "Astrocytes — CAA, IgG",        genes_to_label = astro_candidate_genes)
```

```{r}
#| label: plot-astro-paren-treatment
#| column: page
#| fig-height: 5
#| fig-width: 10
plot_gene_distance_correlation(astro_cor_paren_adu,
  title = "Astrocytes — Parenchymal, Aducanumab", genes_to_label = astro_candidate_genes) |
plot_gene_distance_correlation(astro_cor_paren_igg,
  title = "Astrocytes — Parenchymal, IgG",        genes_to_label = astro_candidate_genes)
```

```{r}
#| label: plot-astro-scatter
#| column: page
#| fig-height: 5
#| fig-width: 12
plot_treatment_scatter(astro_cor_caa_adu,   astro_cor_caa_igg,   "Astrocytes — CAA, Adu vs IgG") |
plot_treatment_scatter(astro_cor_paren_adu, astro_cor_paren_igg, "Astrocytes — Parenchymal, Adu vs IgG") |
plot_treatment_scatter(astro_cor_any_adu,   astro_cor_any_igg,   "Astrocytes — Any plaque, Adu vs IgG")
```

## Venn diagram — astrocyte enriched-near overlap

```{r}
#| label: astro-venn
#| fig-height: 4
#| fig-width: 6
#| fig-cap: "Overlap of genes significantly enriched near plaques (padj < 0.05, rho_flip > 0) in astrocytes, between CAA and parenchymal analyses."

near_caa_astro   <- cor_caa_astro   |> filter(padj < 0.05, rho_flip > 0) |> pull(gene)
near_paren_astro <- cor_paren_astro |> filter(padj < 0.05, rho_flip > 0) |> pull(gene)

ggVennDiagram(
  list(CAA = near_caa_astro, Parenchymal = near_paren_astro),
  label_alpha = 0, label = "count", shape_id = "201"
) +
  labs(title = "Astrocytes — enriched-near genes (padj < 0.05)") +
  theme(legend.position = "none") +
  scale_fill_distiller(palette = "RdBu")

cat("Common enriched-near astrocyte genes (CAA and parenchymal):\n",
    paste(intersect(near_caa_astro, near_paren_astro), collapse = ", "))
```

# Oligodendrocyte gene-distance correlations

Candidate genes from DEG_consolidated model (Layer 3 — Remodeling):
- Near plaque: Abca1, Abcg1, Srebf1, Apod, Aspa, Mbp, Plp1
- Far from plaque: Got1, Ldhb, Mdh2, Hk1 (metabolic downregulation)

```{r}
#| label: run-oligo-correlations
#| eval: false
oligo_candidate_genes <- c(
  "Abca1", "Abcg1", "Srebf1", "Apod", "Aspa",
  "Mbp", "Plp1", "Got1", "Ldhb", "Mdh2", "Hk1"
)

cor_caa_oligo   <- compute_gene_distance_correlation(
  slides_oligo, dist_oligo, distance_col = "dist_caa",   min_expressing_cells = 50)
cor_paren_oligo <- compute_gene_distance_correlation(
  slides_oligo, dist_oligo, distance_col = "dist_paren", min_expressing_cells = 50)
cor_any_oligo   <- compute_gene_distance_correlation(
  slides_oligo, dist_oligo, distance_col = "dist_any",   min_expressing_cells = 50)

# qs2::qs_save(cor_caa_oligo,   "spatial/oligo_cor_caa.qs2")
# qs2::qs_save(cor_paren_oligo, "spatial/oligo_cor_paren.qs2")
# qs2::qs_save(cor_any_oligo,   "spatial/oligo_cor_any.qs2")
```

```{r}
#| label: run-oligo-correlations-by-treatment
#| eval: false
dist_oligo_adu <- dist_oligo |> filter(Treatment.Group == "Adu")
dist_oligo_igg <- dist_oligo |> filter(Treatment.Group == "IgG")

for (trt in c("adu", "igg")) {
  dist_trt <- if (trt == "adu") dist_oligo_adu else dist_oligo_igg
  for (dcol in c("dist_caa", "dist_paren", "dist_any")) {
    res    <- compute_gene_distance_correlation(slides_oligo, dist_trt,
                distance_col = dcol, min_expressing_cells = 50)
    suffix <- sub("dist_", "", dcol)
    qs2::qs_save(res, paste0("spatial/oligo_cor_", suffix, "_", trt, ".qs2"))
  }
}
```

```{r}
#| label: load-oligo-correlations
oligo_candidate_genes <- c(
  "Abca1", "Abcg1", "Srebf1", "Apod", "Aspa",
  "Mbp", "Plp1", "Got1", "Ldhb", "Mdh2", "Hk1"
)

cor_caa_oligo   <- qs2::qs_read("spatial/oligo_cor_caa.qs2")
cor_paren_oligo <- qs2::qs_read("spatial/oligo_cor_paren.qs2")
cor_any_oligo   <- qs2::qs_read("spatial/oligo_cor_any.qs2")

oligo_cor_caa_adu   <- qs2::qs_read("spatial/oligo_cor_caa_adu.qs2")
oligo_cor_caa_igg   <- qs2::qs_read("spatial/oligo_cor_caa_igg.qs2")
oligo_cor_paren_adu <- qs2::qs_read("spatial/oligo_cor_paren_adu.qs2")
oligo_cor_paren_igg <- qs2::qs_read("spatial/oligo_cor_paren_igg.qs2")
oligo_cor_any_adu   <- qs2::qs_read("spatial/oligo_cor_any_adu.qs2")
oligo_cor_any_igg   <- qs2::qs_read("spatial/oligo_cor_any_igg.qs2")
```

## Rank plots — oligodendrocytes

```{r}
#| label: plot-oligo-caa-paren
#| column: page
#| fig-height: 5
#| fig-width: 10
#| fig-cap: "Spearman gene-distance correlation for oligodendrocytes. Top 5 enriched-near and enriched-far genes are automatically labeled; candidate Layer 3 genes (Abca1, Abcg1, Srebf1, Apod, Got1, Ldhb) are additionally labeled if present."

p_caa_oligo   <- plot_gene_distance_correlation(
  cor_caa_oligo,
  title          = "Oligodendrocytes — CAA",
  genes_to_label = oligo_candidate_genes
)

p_paren_oligo <- plot_gene_distance_correlation(
  cor_paren_oligo,
  title          = "Oligodendrocytes — Parenchymal",
  genes_to_label = oligo_candidate_genes
)

(p_caa_oligo | p_paren_oligo) +
  plot_annotation(title = "Oligodendrocyte gene-distance correlation") +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

## Top oligodendrocyte genes

```{r}
#| label: oligo-top-tables
show_top(cor_caa_oligo,   "Top genes — oligodendrocytes, distance to CAA")
show_top(cor_paren_oligo, "Top genes — oligodendrocytes, distance to parenchymal")
```

## Treatment-split rank plots — oligodendrocytes

```{r}
#| label: plot-oligo-caa-treatment
#| column: page
#| fig-height: 5
#| fig-width: 10
plot_gene_distance_correlation(oligo_cor_caa_adu,
  title = "Oligodendrocytes — CAA, Aducanumab", genes_to_label = oligo_candidate_genes) |
plot_gene_distance_correlation(oligo_cor_caa_igg,
  title = "Oligodendrocytes — CAA, IgG",        genes_to_label = oligo_candidate_genes)
```

```{r}
#| label: plot-oligo-paren-treatment
#| column: page
#| fig-height: 5
#| fig-width: 10
plot_gene_distance_correlation(oligo_cor_paren_adu,
  title = "Oligodendrocytes — Parenchymal, Aducanumab", genes_to_label = oligo_candidate_genes) |
plot_gene_distance_correlation(oligo_cor_paren_igg,
  title = "Oligodendrocytes — Parenchymal, IgG",        genes_to_label = oligo_candidate_genes)
```

```{r}
#| label: plot-oligo-scatter
#| column: page
#| fig-height: 5
#| fig-width: 12
plot_treatment_scatter(oligo_cor_caa_adu,   oligo_cor_caa_igg,   "Oligodendrocytes — CAA, Adu vs IgG") |
plot_treatment_scatter(oligo_cor_paren_adu, oligo_cor_paren_igg, "Oligodendrocytes — Parenchymal, Adu vs IgG") |
plot_treatment_scatter(oligo_cor_any_adu,   oligo_cor_any_igg,   "Oligodendrocytes — Any plaque, Adu vs IgG")
```

## Venn diagram — oligodendrocyte enriched-near overlap

```{r}
#| label: oligo-venn
#| fig-height: 4
#| fig-width: 6
#| fig-cap: "Overlap of genes significantly enriched near plaques (padj < 0.05, rho_flip > 0) in oligodendrocytes."

near_caa_oligo   <- cor_caa_oligo   |> filter(padj < 0.05, rho_flip > 0) |> pull(gene)
near_paren_oligo <- cor_paren_oligo |> filter(padj < 0.05, rho_flip > 0) |> pull(gene)

ggVennDiagram(
  list(CAA = near_caa_oligo, Parenchymal = near_paren_oligo),
  label_alpha = 0, label = "count", shape_id = "201"
) +
  labs(title = "Oligodendrocytes — enriched-near genes (padj < 0.05)") +
  theme(legend.position = "none") +
  scale_fill_distiller(palette = "RdBu")

cat("Common enriched-near oligodendrocyte genes (CAA and parenchymal):\n",
    paste(intersect(near_caa_oligo, near_paren_oligo), collapse = ", "))
```

# Cross-cell-type comparison

## Astrocytes vs oligodendrocytes — candidate gene heatmap

Visualise the rho_flip values for all DEG_consolidated candidate genes across both
cell types and both plaque types in a single heatmap.

```{r}
#| label: candidate-gene-heatmap
#| fig-height: 6
#| fig-width: 8
#| fig-cap: "Spearman rho_flip for candidate Layer 2 (astrocyte) and Layer 3 (oligodendrocyte) genes across CAA and parenchymal distance. Positive (orange) = enriched near plaque; negative (purple) = enriched far from plaque. Grey fill = gene not tested (< 50 expressing cells)."

all_candidates <- union(astro_candidate_genes, oligo_candidate_genes)

heatmap_df <- bind_rows(
  cor_caa_astro   |> mutate(cell_type = "Astrocytes",      plaque = "CAA"),
  cor_paren_astro |> mutate(cell_type = "Astrocytes",      plaque = "Parenchymal"),
  cor_caa_oligo   |> mutate(cell_type = "Oligodendrocytes", plaque = "CAA"),
  cor_paren_oligo |> mutate(cell_type = "Oligodendrocytes", plaque = "Parenchymal")
) |>
  filter(gene %in% all_candidates) |>
  mutate(
    facet = paste(cell_type, plaque, sep = " — "),
    sig_label = case_when(
      padj < 0.001 ~ "***",
      padj < 0.01  ~ "**",
      padj < 0.05  ~ "*",
      TRUE         ~ ""
    )
  )

ggplot(heatmap_df, aes(x = facet, y = gene, fill = rho_flip)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = sig_label), size = 4, vjust = 0.75) +
  scale_fill_gradient2(
    low  = "#7B68AE", mid = "white", high = "#E07B39",
    midpoint = 0,
    name = "rho_flip\n(+ = near plaque)"
  ) +
  labs(
    title = "Candidate gene-distance correlations",
    subtitle = "* padj < 0.05  ** padj < 0.01  *** padj < 0.001",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```
