---
title: "Astrocyte and oligodendrocyte proximal treatment DE"
subtitle: " Adu vs IgG in parenchymal- and CAA-plaque-proximal glia DEGs"
date: last-modified
---

# Rationale

Notebook 8 showed that microglia near parenchymal plaques carry a nominally
significant treatment signal (Ccl12, Ifitm3, Nlrp3 at p < 0.01). Here we extend
the same pseudobulk DESeq2 framework to **astrocytes** and **oligodendrocytes** to
test whether the amplification and remodeling layers of the three-layer model also
show Adu vs IgG differences near plaques.

- **Astrocytes** (Layer 2 — Amplification): hypothesised up: Ifitm3, Ccl2, Agt
- **Oligodendrocytes** (Layer 3 — Remodeling): hypothesised up: Abca1, Srebf1, Apod;
  hypothesised down: Got1, Ldhb

Pseudobulk aggregation per brain (n = 3 Adu vs n = 3 IgG) is used throughout. No
genes are expected to survive BH correction; we report nominal p < 0.01 as
exploratory candidates, consistent with notebook 8.

# Setup and data loading

```{r}
#| label: setup
source("R/setup.R")
library(DESeq2)
library(EnhancedVolcano)
library(patchwork)
library(gridExtra)
library(ggridges)

data <- load_slides_and_distances("full")
slide1 <- data$slide1; slide2 <- data$slide2
cell_distances_1 <- data$distances_1; cell_distances_2 <- data$distances_2

## Identify cell-types of interest

ct_astro <- "Astrocyte"
ct_oligo <- "Oligodendrocyte"
ct_opc   <- "OPC"
```





## Add proximity metadata to full objects

```{r}
#| label: add-proximity-metadata
add_proximity <- function(seurat_obj, cell_distances) {
  prox <- cell_distances |>
    mutate(
      prox_caa   = if_else(dist_caa < 50,   "proximal", "distal"),
      prox_paren = if_else(dist_paren < 50,  "proximal", "distal"),
      prox_any   = if_else(pmin(dist_caa, dist_paren) < 50, "proximal", "distal")
    ) |>
    column_to_rownames("cell") |>
    select(dist_caa, dist_paren, prox_caa, prox_paren, prox_any)

  AddMetaData(seurat_obj, prox)
}

slide1 <- add_proximity(slide1, cell_distances_1)
slide2 <- add_proximity(slide2, cell_distances_2)
```

## Subset to target cell types

```{r}
#| label: subset-cell-types
# Subset to astrocytes
slide1_astro <- subset(slide1, annotatedclusters %in% ct_astro)
slide2_astro <- subset(slide2, annotatedclusters %in% ct_astro)

# Subset to oligodendrocytes
slide1_oligo <- subset(slide1, annotatedclusters %in% ct_oligo)
slide2_oligo <- subset(slide2, annotatedclusters %in% ct_oligo)

message(sprintf("Astrocytes:        slide1 = %d, slide2 = %d",
                ncol(slide1_astro), ncol(slide2_astro)))
message(sprintf("Oligodendrocytes:  slide1 = %d, slide2 = %d",
                ncol(slide1_oligo), ncol(slide2_oligo)))
```

## Cell counts per proximity group

```{r}
#| label: cell-counts
count_proximity <- function(slide_list, label) {
  bind_rows(map(slide_list, \(obj) obj@meta.data |> as_tibble(rownames = "cell"))) |>
    group_by(sample_ID, prox_any) |>
    summarise(n = n(), .groups = "drop") |>
    pivot_wider(names_from = prox_any, values_from = n, values_fill = 0L) |>
    mutate(total = proximal + distal, cell_type = label)
}

bind_rows(
  count_proximity(list(slide1_astro, slide2_astro), "Astrocytes"),
  count_proximity(list(slide1_oligo, slide2_oligo), "Oligodendrocytes")
) |>
  kbl(caption = "Cells per sample — proximity to any plaque (50 um threshold)") |>
  kable_styling("striped", full_width = FALSE)
```

## Ridge plots: distance distributions by treatment

```{r}
#| label: ridge-distances
#| fig-width: 14
#| fig-height: 8
#| fig-cap: "Distance distributions for astrocytes (top) and oligodendrocytes (bottom) by treatment group. Vertical dashed line = 50 um proximal threshold. Distances capped at 500 um."

build_ridge_df <- function(slide_list, label) {
  bind_rows(map(slide_list, \(obj) {
    obj@meta.data |>
      as_tibble(rownames = "cell") |>
      select(cell, sample_ID, Treatment.Group, dist_caa, dist_paren)
  })) |>
    pivot_longer(
      cols = c(dist_caa, dist_paren),
      names_to = "plaque_type", values_to = "distance"
    ) |>
    mutate(
      plaque_type   = recode(plaque_type, dist_caa = "CAA", dist_paren = "Parenchymal"),
      distance_cap  = pmin(distance, 500),
      cell_type     = label
    )
}

ridge_df <- bind_rows(
  build_ridge_df(list(slide1_astro, slide2_astro), "Astrocytes"),
  build_ridge_df(list(slide1_oligo, slide2_oligo), "Oligodendrocytes")
)

ggplot(ridge_df, aes(x = distance_cap, y = Treatment.Group, fill = Treatment.Group)) +
  geom_density_ridges(alpha = 0.6, scale = 1.2, rel_min_height = 0.01) +
  geom_vline(xintercept = 50, linetype = "dashed", colour = "grey30") +
  facet_grid(cell_type ~ plaque_type) +
  scale_fill_manual(values = c("IgG" = "#4DAF4A", "Adu" = "#E41A1C")) +
  labs(
    x    = "Distance to nearest plaque (um)",
    y    = NULL,
    fill = "Treatment"
  ) +
  theme_minimal(base_size = 15) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 14))
```

# Pseudobulk helpers

Functions are identical to notebook 8: aggregate counts per sample within a proximity
group, then run DESeq2 Adu vs IgG.

```{r}
#| label: define-pseudobulk-builder
build_pseudobulk <- function(slide_list, group_col) {
  all_counts <- map(slide_list, \(obj) {
    LayerData(obj, assay = DefaultAssay(obj), layer = "counts")
  })
  all_meta <- map(slide_list, \(obj) {
    obj@meta.data |>
      rownames_to_column("cell") |>
      mutate(pb_group = .data[[group_col]])
  })

  shared_genes <- Reduce(intersect, map(all_counts, rownames))
  counts <- do.call(cbind, map(all_counts, \(m) m[shared_genes, ]))
  meta   <- bind_rows(all_meta)

  groups <- unique(meta$pb_group)
  indicator <- Matrix::sparseMatrix(
    i = match(meta$cell, colnames(counts)),
    j = match(meta$pb_group, groups),
    x = 1,
    dims = c(ncol(counts), length(groups)),
    dimnames = list(colnames(counts), groups)
  )

  list(
    pb_counts = as.matrix(counts %*% indicator),
    meta      = meta,
    groups    = groups
  )
}
```

```{r}
#| label: define-deseq2-runner
run_deseq2 <- function(
    pb_counts,
    col_data,
    design_formula,
    contrast,
    plot_title    = "DE results",
    plot_subtitle = "",
    pval_threshold = 0.01
) {
  dds <- DESeqDataSetFromMatrix(
    countData = pb_counts,
    colData   = col_data,
    design    = design_formula
  )
  keep <- rowSums(counts(dds)) >= 10
  dds  <- dds[keep, ]
  dds  <- DESeq(dds)
  res  <- results(dds, contrast = contrast)

  res_df <- as.data.frame(res) |>
    rownames_to_column("gene") |>
    as_tibble() |>
    dplyr::rename(p_val = pvalue, p_val_adj = padj, avg_log2FC = log2FoldChange) |>
    arrange(p_val_adj)

  sig_df    <- res_df |> filter(p_val < pval_threshold)
  ev_df     <- res_df |> column_to_rownames("gene")
  sig_genes <- sig_df$gene

  keyvals <- rep("grey60", nrow(ev_df))
  names(keyvals) <- rownames(ev_df)
  keyvals[sig_genes] <- "red2"

  fc_range  <- range(ev_df$avg_log2FC, na.rm = TRUE)
  xlim_vals <- c(floor(fc_range[1]), ceiling(fc_range[2]))

  volcano <- EnhancedVolcano(
    ev_df,
    lab            = rownames(ev_df),
    selectLab      = sig_genes,
    x              = "avg_log2FC",
    y              = "p_val",
    ylim           = c(0, 3),
    title          = plot_title,
    subtitle       = plot_subtitle,
    labFace        = "italic",
    labSize        = 6,
    caption        = "",
    pCutoff        = 1,
    FCcutoff       = 0.001,
    colCustom      = keyvals,
    legendPosition = "none",
    drawConnectors = TRUE,
    directionConnectors = "both",
    max.overlaps   = Inf,
    xlim           = xlim_vals
  )

  if (length(sig_genes) > 0) {
    raw_boundary <- max(ev_df[sig_genes, "p_val"], na.rm = TRUE)
    volcano <- volcano +
      geom_hline(yintercept = -log10(raw_boundary),
                 linetype = "dashed", colour = "black")
  }

  list(
    all_results         = ev_df,
    significant_results = sig_df |> column_to_rownames("gene"),
    volcano_plot        = volcano,
    sig_genes           = sig_genes,
    summary = list(total_genes = nrow(ev_df), significant_genes = nrow(sig_df))
  )
}
```

```{r}
#| label: define-treatment-de
run_treatment_de <- function(slide_list, proximity_col, proximity_value, plot_title) {
  slide_list_sub <- map(slide_list, \(obj) {
    subset(obj, cells = colnames(obj)[obj[[proximity_col, drop = TRUE]] == proximity_value])
  })

  pb <- build_pseudobulk(slide_list_sub, "sample_ID")

  treatment_map <- bind_rows(map(slide_list, \(obj) {
    obj@meta.data |> distinct(sample_ID, Treatment.Group)
  })) |> distinct()

  col_data <- data.frame(
    treatment = factor(
      treatment_map$Treatment.Group[match(pb$groups, treatment_map$sample_ID)],
      levels = c("IgG", "Adu")
    ),
    row.names = pb$groups
  )

  run_deseq2(pb$pb_counts, col_data,
             design_formula = ~ treatment,
             contrast       = c("treatment", "Adu", "IgG"),
             plot_title     = plot_title,
             plot_subtitle  = "Adu vs IgG")
}
```

# Gene filtering funnel

```{r}
#| label: gene-funnel
count_after_pb <- function(slide_list, proximity_col, proximity_value) {
  sub <- map(slide_list, \(obj) {
    subset(obj, cells = colnames(obj)[obj[[proximity_col, drop = TRUE]] == proximity_value])
  })
  pb <- build_pseudobulk(sub, "sample_ID")
  n_cells        <- sum(map_int(sub, ncol))
  n_genes_before <- nrow(pb$pb_counts)
  n_genes_after  <- sum(rowSums(pb$pb_counts) >= 10)
  tibble(n_cells = n_cells, genes_in_pb = n_genes_before, genes_after_filter = n_genes_after)
}

panel_genes <- intersect(
  rownames(LayerData(slide1, layer = "counts")),
  rownames(LayerData(slide2, layer = "counts"))
)

astro_caa_stats   <- count_after_pb(list(slide1_astro, slide2_astro), "prox_caa",   "proximal")
astro_paren_stats <- count_after_pb(list(slide1_astro, slide2_astro), "prox_paren", "proximal")
oligo_caa_stats   <- count_after_pb(list(slide1_oligo, slide2_oligo), "prox_caa",   "proximal")
oligo_paren_stats <- count_after_pb(list(slide1_oligo, slide2_oligo), "prox_paren", "proximal")

tribble(
  ~step,                                                   ~count,
  "Xenium panel (shared across slides)",                   length(panel_genes),
  "Astro CAA-proximal: cells",                             astro_caa_stats$n_cells,
  "Astro CAA-proximal: genes after >=10 counts filter",    astro_caa_stats$genes_after_filter,
  "Astro Paren-proximal: cells",                           astro_paren_stats$n_cells,
  "Astro Paren-proximal: genes after >=10 counts filter",  astro_paren_stats$genes_after_filter,
  "Oligo CAA-proximal: cells",                             oligo_caa_stats$n_cells,
  "Oligo CAA-proximal: genes after >=10 counts filter",    oligo_caa_stats$genes_after_filter,
  "Oligo Paren-proximal: cells",                           oligo_paren_stats$n_cells,
  "Oligo Paren-proximal: genes after >=10 counts filter",  oligo_paren_stats$genes_after_filter
) |>
  kbl(caption = "Gene filtering funnel — cells and genes available per group") |>
  kable_styling("striped", full_width = FALSE)
```

# Astrocyte treatment DE

## CAA-proximal astrocytes: Adu vs IgG

```{r}
#| label: astro-trt-caa
#| fig-width: 14
#| fig-height: 6
trt_caa_astro <- run_treatment_de(
  list(slide1_astro, slide2_astro), "prox_caa", "proximal",
  "Astrocytes proximal to CAA — Adu vs IgG"
)

sig_tbl <- if (nrow(trt_caa_astro$significant_results) > 0) {
  trt_caa_astro$significant_results |>
    select(p_val, avg_log2FC, p_val_adj) |>
    mutate(across(where(is.numeric), \(x) signif(x, 3))) |>
    arrange(p_val_adj) |>
    tableGrob(theme = ttheme_minimal(base_size = 14))
} else {
  tableGrob(data.frame(Result = "No significant DEGs"), theme = ttheme_minimal(base_size = 16))
}

trt_caa_astro$volcano_plot + wrap_elements(sig_tbl) + plot_layout(widths = c(2, 1))
```

## Parenchymal-proximal astrocytes: Adu vs IgG

```{r}
#| label: astro-trt-paren
#| fig-width: 14
#| fig-height: 6
trt_paren_astro <- run_treatment_de(
  list(slide1_astro, slide2_astro), "prox_paren", "proximal",
  "Astrocytes proximal to parenchymal — Adu vs IgG"
)

sig_tbl <- if (nrow(trt_paren_astro$significant_results) > 0) {
  trt_paren_astro$significant_results |>
    select(p_val, avg_log2FC, p_val_adj) |>
    mutate(across(where(is.numeric), \(x) signif(x, 3))) |>
    arrange(p_val_adj) |>
    tableGrob(theme = ttheme_minimal(base_size = 14))
} else {
  tableGrob(data.frame(Result = "No significant DEGs"), theme = ttheme_minimal(base_size = 16))
}

trt_paren_astro$volcano_plot + wrap_elements(sig_tbl) + plot_layout(widths = c(2, 1))
```

# Oligodendrocyte treatment DE

## CAA-proximal oligodendrocytes: Adu vs IgG

```{r}
#| label: oligo-trt-caa
#| fig-width: 14
#| fig-height: 6
trt_caa_oligo <- run_treatment_de(
  list(slide1_oligo, slide2_oligo), "prox_caa", "proximal",
  "Oligodendrocytes proximal to CAA — Adu vs IgG"
)

sig_tbl <- if (nrow(trt_caa_oligo$significant_results) > 0) {
  trt_caa_oligo$significant_results |>
    select(p_val, avg_log2FC, p_val_adj) |>
    mutate(across(where(is.numeric), \(x) signif(x, 3))) |>
    arrange(p_val_adj) |>
    tableGrob(theme = ttheme_minimal(base_size = 14))
} else {
  tableGrob(data.frame(Result = "No significant DEGs"), theme = ttheme_minimal(base_size = 16))
}

trt_caa_oligo$volcano_plot + wrap_elements(sig_tbl) + plot_layout(widths = c(2, 1))
```

## Parenchymal-proximal oligodendrocytes: Adu vs IgG

```{r}
#| label: oligo-trt-paren
#| fig-width: 14
#| fig-height: 6
trt_paren_oligo <- run_treatment_de(
  list(slide1_oligo, slide2_oligo), "prox_paren", "proximal",
  "Oligodendrocytes proximal to parenchymal — Adu vs IgG"
)

sig_tbl <- if (nrow(trt_paren_oligo$significant_results) > 0) {
  trt_paren_oligo$significant_results |>
    select(p_val, avg_log2FC, p_val_adj) |>
    mutate(across(where(is.numeric), \(x) signif(x, 3))) |>
    arrange(p_val_adj) |>
    tableGrob(theme = ttheme_minimal(base_size = 14))
} else {
  tableGrob(data.frame(Result = "No significant DEGs"), theme = ttheme_minimal(base_size = 16))
}

trt_paren_oligo$volcano_plot + wrap_elements(sig_tbl) + plot_layout(widths = c(2, 1))
```

# Summary: candidate Layer 2 and 3 genes

```{r}
#| label: summary-table
# Pull all results and check candidate genes
layer2_candidates <- c("Ifitm3", "Ccl2", "Ccl12", "Cxcl10", "Agt")
layer3_candidates <- c("Abca1", "Abcg1", "Srebf1", "Apod", "Got1", "Ldhb")

extract_candidates <- function(de_result, candidates, label) {
  de_result$all_results |>
    as_tibble(rownames = "gene") |>
    filter(gene %in% candidates) |>
    mutate(
      comparison = label,
      sig        = p_val < 0.01
    ) |>
    select(gene, avg_log2FC, p_val, p_val_adj, comparison, sig) |>
    arrange(p_val)
}

bind_rows(
  extract_candidates(trt_caa_astro,   layer2_candidates, "Astro — CAA proximal"),
  extract_candidates(trt_paren_astro, layer2_candidates, "Astro — Paren proximal"),
  extract_candidates(trt_caa_oligo,   layer3_candidates, "Oligo — CAA proximal"),
  extract_candidates(trt_paren_oligo, layer3_candidates, "Oligo — Paren proximal")
) |>
  mutate(across(c(avg_log2FC), \(x) round(x, 3)),
         across(c(p_val, p_val_adj), \(x) formatC(x, format = "e", digits = 2))) |>
  kbl(caption = "Candidate Layer 2/3 genes across all comparisons (nominal p < 0.01 = TRUE)") |>
  kable_styling("striped", full_width = FALSE) |>
  row_spec(which(bind_rows(
    extract_candidates(trt_caa_astro,   layer2_candidates, "Astro — CAA proximal"),
    extract_candidates(trt_paren_astro, layer2_candidates, "Astro — Paren proximal"),
    extract_candidates(trt_caa_oligo,   layer3_candidates, "Oligo — CAA proximal"),
    extract_candidates(trt_paren_oligo, layer3_candidates, "Oligo — Paren proximal")
  )$sig), bold = TRUE, color = "red")
```
