---
title: "Tissue domain characterization (I)"
subtitle: "BANSKY segmentation (lambda = 0.8, SCT normalized data)"
date: last-modified
---

**BANKSY (Building Aggregates with a Neighborhood Kernel and Spatial Yardstick)** augments each cell's own transcriptome with the weighted mean expression of its spatial neighbours. At `lambda = 0.8`, neighbourhood context dominates, grouping cells into contiguous **tissue domains** rather than transcriptomic cell types.

Reference: [Singhal et al., *Nature Genetics*, 2024](https://doi.org/10.1038/s41588-024-01664-3)

# Setup and data loading

`aria` is the full merged Seurat object (6 brains, 358 399 cells total) after Chloe correction. Assay `SCT_SPLIT` contains SPLIT-denoised, SCTransformed counts (474 features) for 346 928 cells that passed the SPLIT quality filter. BANKSY is run on this filtered cell set. Spatial coordinates are in metadata columns `x` and `y`.

```{r}
#| label: setup

# BiocManager::install('Banksy')
# remotes::install_github('satijalab/seurat-wrappers')

source("R/setup.R")
library(Banksy)
library(SeuratWrappers)
library(patchwork)
library(Polychrome)


```

# Method: BANKSY at lambda = 0.8, use_agf = TRUE

BANKSY constructs an augmented expression matrix by concatenating:

1. The cell's own (normalised) gene expression.
2. Two spatial kernels:
  - **Weighted Neighborhood Mean**: average gene expression of physical neighbors, weighted by distance (Gaussian envelope)
  - **Azimuthal Gabor Filter (AGF)**: estimates the "spatial texture" or gradient of gene expression within the cellular neighborhood (_not used by default !!??_)

The mixing parameter **`lambda = 0.8`** places 80 % weight on the neighbourhood context, activating **tissue domain segmentation mode**. Cells are embedded in this augmented space, dimensionality-reduced with PCA, and clustered with Leiden.

The 6 brains are spread across **two slides**. Within each slide, the 3 brains are placed at different physical positions so their `x`/`y` coordinates are already non-overlapping. Between slides, however, each slide has its own coordinate system starting from (0, 0), so slide 1 and slide 2 coordinates do overlap. We therefore set `group = "Slide"` to stagger coordinates between slides only, preserving the real within-slide spatial layout. `split.scale = TRUE` applies within-slide z-scaling before concatenation, accounting for minor technical differences between imaging runs.

> **Important**: do not call `ScaleData()` on the resulting BANKSY assay. `RunBanksy` already populates `scale.data` with the lambda-weighted features. Re-scaling would erase the neighbourhood weighting.

# Run BANKSY

```{r}
#| label: run-banksy
#| eval: false

# Load full object
aria <- qs2::qs_read(paste0(seurat_path, "20260320_fullobj_correctedChloe.qs2"))

# Load cleaned microglia object
microglia <- qs2::qs_read(paste0(seurat_path, "20260323_updated_mg_kai_correctedChloe.qs2"))

# Keep only the microglia subtype labels from the metadata of the cleaned microglia-only object.
mg_subcluster <- microglia@meta.data |> 
  select(sub.cluster)


DefaultAssay(aria) <- "SCT_SPLIT"

## Subset cells in the SCT_SPLIT assay of the full object
# SCT_SPLIT exists only for SPLIT-passing cells.
# Why: BANKSY uses assay matrices directly; restricting to Cells(SCT_SPLIT) prevents
# downstream failures from trying to run spatial modeling on cells without SCT_SPLIT data.
aria_sct_split <- subset(aria, cells = Cells(aria[["SCT_SPLIT"]]))

# Transfer subtype labels from cleaned microglia object into full object.
aria_sct_split$microglia_sub.cluster <- mg_subcluster[colnames(aria_sct_split), "sub.cluster"]


aria_sct_split$annotatedclusters_finer <- if_else(is.na(aria_sct_split$microglia_sub.cluster), aria_sct_split$annotatedclusters, aria_sct_split$microglia_sub.cluster)



# create new variable "annotatedclusters_finer" in the metadata including major cell type and microglia subtype
aria_sct_split$annotatedclusters_finer <- ifelse(
  is.na(aria_sct_split$microglia_sub.cluster),
  as.character(aria_sct_split$annotatedclusters),
  as.character(aria_sct_split$microglia_sub.cluster))
# Cells not present in cleaned microglia return "Microglia" in "annotatedclusters_finer".
# This is expected because 11,796 microglia were removed during microglia QC cleaning.

# Drop cells still labeled exactly "Microglia" (i.e., no refined subtype assigned).
aria_sct_split <-subset(aria_sct_split, subset = annotatedclusters_finer != "Microglia")

# qs2::qs_save(aria_sct_split, paste0(seurat_path, "20260402_fullobj_correctedChloe_sct_SPLIT_microglia_cleaned.qs2"))

aria_sct_split <- RunBanksy(
  aria_sct_split,
  lambda      = 0.8,
  assay       = "SCT_SPLIT",
  slot        = "data",        # SCT-normalised expression
  features    = "all",    # 474 variable features from SCT_SPLIT
  k_geom      = 15,            # spatial neighbours per cell
  dimx        = "x",           # x coordinate in metadata
  dimy        = "y",           # y coordinate in metadata
  use_agf     = TRUE,          # use AGF (false by default)
  group       = "Slide",       # stagger coordinates between slides (overlap between slides, not within)
  split.scale = TRUE)           # within-slide z-scaling)

# Active assay is now BANKSY; do NOT call ScaleData
aria_sct_split <- RunPCA(aria_sct_split,
               assay    = "BANKSY",
               features = rownames(aria_sct_split),  # all BANKSY features
               npcs     = 50)

aria_sct_split <- RunUMAP(aria_sct_split, dims = 1:30, seed.use = 42,
                reduction.name = "umap_banksy", reduction.key = "BKSYUMAP_")

aria_sct_split <- FindNeighbors(aria_sct_split, dims = 1:30)
aria_sct_split <- FindClusters(aria_sct_split, resolution = 0.5, random.seed = 42)

# Store BANKSY domains in an explicit column (avoids confusion with pre-existing banksy_domain)
aria_sct_split$banksy_domain <- aria_sct_split$BANKSY_snn_res.0.5
Idents(aria_sct_split) <- "banksy_domain"

qs2::qs_save(aria_sct_split, paste0(seurat_path, "20260402_fullobj_correctedChloe_sct_SPLIT_microglia_cleaned.qs2_banksy_agf=TRUE_sct.qs2"))
```

```{r}
#| label: load-banksy


aria_sct_split <- qs2::qs_read(paste0(seurat_path, "20260402_fullobj_correctedChloe_sct_SPLIT_microglia_cleaned.qs2_banksy_agf=TRUE_sct.qs2"))


```

# UMAP

```{r}
#| label: umap-domains
DimPlot(aria_sct_split, reduction = "umap_banksy",
        group.by  = "banksy_domain",
        label     = TRUE,
        label.size = 4,
        pt.size   = 0.3) + 
  labs(title = "BANKSY tissue domains",
        subtitle = "lambda = 0.8 / resolution = 0.5", colour = "Domain") +
  theme_minimal(base_size = 13)+
  theme(legend.position = "none")
```



```{r}
#| label: umap-treatment
#| column: page
#| fig-width: 10
DimPlot(aria_sct_split, reduction = "umap_banksy",
        split.by = "Treatment.Group",
        pt.size  = 0.3) 

  labs(title = "Treatment group on BANKSY UMAP") +
  theme_minimal(base_size = 13)
```

# Spatial domain maps

Domains plotted in tissue space using the `x`/`y` metadata coordinates, one panel per brain. The staggered-coordinate overview (all brains side-by-side) uses `staggered_x`/`staggered_y` written by `RunBanksy`.

## Staggered overview (all 6 brains)

```{r}
#| label: spatial-staggered
#| column: page
#| fig-height: 8
#| fig-width: 12
FeatureScatter(
  aria_sct_split,
  feature1 = "staggered_sdimx",
  feature2 = "staggered_sdimy",
  group.by = "banksy_domain",
  pt.size  = 0.8
) +
  labs(title  = "BANKSY domains — all brains (staggered)",
       x      = "Staggered x (µm)",
       y      = "Staggered y (µm)",
       colour = "Domain") +
  theme_minimal(base_size = 12)
```

## Per-brain spatial maps

```{r}
#| label: spatial-domains
#| column: page
#| fig-height: 8
#| fig-width: 12

meta <- aria_sct_split@meta.data |> as_tibble(rownames = "cell")

# Colour palette: one colour per domain
n_domains   <- length(unique(aria_sct_split$banksy_domain))
domain_cols <- kelly.colors(n_domains) |> setNames(sort(unique(aria_sct_split$banksy_domain)))

sample_ids <- sort(unique(meta$sample_ID))

plots <- map2(sample_ids, seq_along(sample_ids), \(sid, i) {
  df <- meta |> filter(sample_ID == sid)
  ggplot(df, aes(x = x, y = y, colour = banksy_domain)) +
    geom_point(size = 0.2, alpha = 0.7, show.legend = (i == 1)) +
    scale_colour_manual(
      values = domain_cols,
      limits = names(domain_cols),
      drop = FALSE
    ) +
    coord_equal() +
    scale_y_reverse() +
    labs(title = sid, colour = "Domain") +
    theme_void(base_size = 10) +
    theme(legend.position = if (i == 1) "left" else "none")
})

wrap_plots(plots, nrow = 2) +
  plot_annotation(title = "BANKSY tissue domains per brain") &
  guides(colour = guide_legend(override.aes = list(size = 4)))
```

# Domain composition

Cell-type and treatment composition of each BANKSY domain.

```{r}
#| label: domain-composition
#| column: page
#| fig-height: 5
#| fig-width: 12
ct_frac <- meta |>
  count(banksy_domain, annotatedclusters_finer) |>
  group_by(banksy_domain) |>
  mutate(frac = n / sum(n)) |>
  ungroup()

ggplot(ct_frac, aes(x = banksy_domain, y = frac, fill = annotatedclusters_finer)) +
  geom_col() +
  labs(title = "Cell-type composition per BANKSY domain",
       x     = "Domain",
       y     = "Fraction of cells",
       fill  = "Cell type") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "right",
        axis.text.x     = element_text(angle = 45, hjust = 1))
```

```{r}
#| label: treatment-composition
#| fig-height: 3
#| fig-width: 8
trt_frac <- meta |>
  count(banksy_domain, Treatment.Group) |>
  group_by(banksy_domain) |>
  mutate(frac = n / sum(n)) |>
  ungroup()

ggplot(trt_frac, aes(x = banksy_domain, y = frac, fill = Treatment.Group)) +
  geom_col() +
  scale_fill_manual(values = c(Adu = "#E07B39", IgG = "#7B68AE")) +
  labs(title = "Treatment composition per BANKSY domain",
       x     = "Domain",
       y     = "Fraction of cells",
       fill  = "Treatment") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Domain marker genes

```{r}
#| label: domain-markers
#| eval: false

DefaultAssay(aria_sct_split) <- "SCT_SPLIT"
Idents(aria_sct_split)       <- "banksy_domain"  # explicit BANKSY domains, not seurat_clusters
markers <- FindAllMarkers(
  aria_sct_split,
  assay           = "SCT_SPLIT",
  only.pos        = TRUE,
  min.pct         = 0.1,
  logfc.threshold = 0.25,
  test.use        = "wilcox")
  
write_csv(markers, "spatial/banksy_domain_markers.csv")
```

```{r}
#| label: load-markers
markers <- read_csv("spatial/banksy_domain_markers.csv")
```

```{r}
#| label: top-markers-table
library(kableExtra)

top_markers <- markers |>
  group_by(cluster) |>
  slice_min(p_val_adj, n = 5) |>
  ungroup() |>
  dplyr::select(cluster, gene, avg_log2FC, pct.1, pct.2, p_val_adj)

top_markers |>
  group_by(cluster) |>
  slice_max(avg_log2FC, n = 5) |>
  kbl(caption = "Top 5 marker genes per BANKSY domain") |>
  kable_styling("striped", full_width = FALSE)
```

```{r}
#| label: marker-heatmap
#| fig-height: 12
#| fig-width: 8
top3 <- markers |>
  group_by(cluster) |>
  slice_max(avg_log2FC, n = 3) |>
  pull(gene) |>
  unique()

DotPlot(aria_sct_split,
          assay    = "SCT_SPLIT",
          cols = c("#2f82db","#d84434"),
          features = top3,
          group.by = "banksy_domain") +
  coord_flip()+
  theme(axis.text.y = element_text(size = 10))
```
