---
title: "Co-registration — Akhil (single annotator)"
subtitle: "ThioS quantification and HALO analysis"
date: last-modified
execute:
  echo: true
  warning: false
embed-resources: true
format: html
code-fold: true
tags: [xenium, coregistration, halo]
---

```{r}
library(readxl)
library(tidyverse)
library(kableExtra)
library(Seurat)
library(RANN)
library(compareGroups)
options(scipen = 999)

# ── CAA labels (all 6 brains in one file, sheets "Brain1_AP" … "Brain6_AP") ──
caa_path <- "HALO_thioS_Akhil/20260217_CAAquant_AP.xlsx"

caa_akhil <- excel_sheets(caa_path) |>
  set_names() |>
  map(\(sheet) read_excel(caa_path, sheet = sheet)) |>
  bind_rows(.id = "sheet") |>
  mutate(
    brain = str_remove(sheet, "_AP"),        # "Brain1" … "Brain6"
    index = paste0(tolower(brain), "_", Id)  # normalise case; Id matches Object Id in CSV
  ) |>
  pull(index)

# ── Sample dictionary ─────────────────────────────────────────────────────────
sample_dict <- tribble(
  ~brain,   ~sample,   ~treatment, ~slide,
  "Brain1", "KK4_465", "IgG",      1L,
  "Brain2", "KK4_492", "Adu",      1L,
  "Brain3", "KK4_504", "IgG",      1L,
  "brain4", "KK4_496", "IgG",      2L,
  "brain5", "KK4_502", "Adu",      2L,
  "brain6", "KK4_464", "Adu",      2L
)

# ── HALO plaque CSVs ─────────────────────────────────────────────────────────
# Brain naming in "Analysis Region":
#   Slide 1 (0021991): "Brain1", "Brain2", "Brain3"  (capital B)
#   Slide 2 (0021998): "brain4", "brain5", "brain6"  (lowercase b)
# We normalise to lowercase before building the index so it matches caa_akhil.

thios_all <- list.files(
  "HALO_thioS_Akhil",
  pattern    = "^20260220.*AP\\.csv$",
  full.names = TRUE) |>
  map(\(x) read_csv(x, locale = locale(encoding = "latin1"))) |>
  bind_rows() |>
  rename_with(\(x) str_remove(x, "\\s*\\(.*\\)$")) |>
  rename(brain = "Analysis Region", objectID = "Object Id") |>
  mutate(
    index       = paste0(tolower(brain), "_", objectID),
    plaque_type = if_else(index %in% caa_akhil, "CAA", "Parenchymal")
  ) |>
  left_join(sample_dict, by = "brain")

# ── Split by slide ────────────────────────────────────────────────────────────
thios_1 <- thios_all |> filter(slide == 1)
thios_2 <- thios_all |> filter(slide == 2)

# ── Xenium Seurat objects  ──────
path <- "/Users/arbones/Dropbox/SyncBriefcase/LAB/UK/Xenium/ARIA_Akhil/aria_analysis/seurat_objects/"
# full_object <- qs2::qs_read(paste0(path,"20260217_Part2_fullobj.qs2"))
# slide1 <- subset(full_object,subset = Slide == "Slide1")
# slide2 <- subset(full_object,subset = Slide == "Slide2")
# qs2::qs_save(slide1, paste0(path,"20260217_Part2_fullobj_slide1.qs2"))
# qs2::qs_save(slide2, paste0(path,"20260217_Part2_fullobj_slide2.qs2"))
slide1 <- qs2::qs_read(paste0(path,"20260217_Part2_fullobj_slide1.qs2"))
slide2 <- qs2::qs_read(paste0(path,"20260217_Part2_fullobj_slide2.qs2"))
```

# HALO analysis

All annotations performed by Akhil Palleria (single annotator; no inter-observer variability).

## Metrics per sample

Thioflavin-S (ThioS)-positive amyloid plaques were quantified across six brain samples (Brain1–6; IgG control, n=3; Aducanumab-treated, n=3) using HALO image analysis software, with all annotations performed by a single rater (A.P.) to eliminate inter-observer variability.  

Two plaque phenotypes were identified: cerebrovascular (CAA) and parenchymal plaques. Across all brains, parenchymal plaques were substantially more numerous than CAA plaques, with per-brain counts ranging from 2,681 to 3,961 (parenchymal) versus 284 to 446 (CAA). Consistent with this, parenchymal plaques accounted for the majority of total annotated area in every sample.



```{r}
t1 <- bind_rows(thios_1, thios_2) |>
  group_by(brain, sample, plaque_type) |>
  summarise(
    n_plaques  = n(),
    total_area = sum(Area),
    .groups    = "drop"
  )

t1 |> kbl() |> kable_styling("striped")
```

Group-level comparison revealed no statistically significant difference between Aducanumab-treated and IgG-control animals in either total CAA area (41,759 ± 17,904 vs. 39,697 ± 6,076 µm²; p=0.865), total parenchymal plaque area (410,021 ± 121,107 vs. 374,302 ± 40,078 µm²; p=0.668), or the proportion of total plaque burden attributable to CAA (8.97 ± 1.32% vs. 9.69 ± 2.13%; p=0.653).
```{r}
t1_wide <- t1 |>
  pivot_wider(id_cols = c(brain, sample), names_from = plaque_type, values_from = total_area) |>
  mutate(CAA_percent = CAA / (Parenchymal + CAA) * 100) |>
  inner_join(sample_dict, by = c("brain", "sample"))

caa1 <- compareGroups(treatment ~ CAA + Parenchymal + CAA_percent, data = t1_wide)
export2md(createTable(caa1, show.p.overall = T), caption = "")

```

## Plaque centroids

To enable spatial co-registration with single-cell Xenium transcriptomic data, the centroid of each plaque was computed as the midpoint of the HALO bounding box and converted from pixel coordinates to micrometres using a calibration factor of 0.2125 µm/pixel. This coordinate transformation provides a common spatial reference frame for subsequent nearest-neighbour analyses.

Centroid = midpoint of HALO bounding box, converted from pixels to µm (factor = 0.2125).

```{r}
make_centroids <- function(thios) {
  thios |>
    mutate(
      x_plaque = ((XMin + XMax) / 2) * 0.2125,
      y_plaque = ((YMin + YMax) / 2) * 0.2125,
      plaque_id = index
    ) |>
    select(plaque_id, brain, sample, treatment, plaque_type, x_plaque, y_plaque)
}

plaque_centroids_1 <- make_centroids(thios_1)
plaque_centroids_2 <- make_centroids(thios_2)

caa_coords_1   <- plaque_centroids_1 |> filter(plaque_type == "CAA")
caa_coords_2   <- plaque_centroids_2 |> filter(plaque_type == "CAA")
paren_coords_1 <- plaque_centroids_1 |> filter(plaque_type == "Parenchymal")
paren_coords_2 <- plaque_centroids_2 |> filter(plaque_type == "Parenchymal")
```

## Cell-to-plaque distances (k-d tree)

Minimum Euclidean distances from **each Xenium-profiled** cell to its nearest CAA plaque (dist_caa) and nearest parenchymal plaque (dist_paren) were computed using a k-d tree algorithm. The resulting distance distributions were markedly different between plaque types. For CAA plaques, the distribution had a modal distance of approximately 200–300 µm and a long tail extending to ~1,500 µm, reflecting the relatively sparse and focal distribution of vascular amyloid deposits. By contrast, the majority of cells were within ~50 µm of a parenchymal plaque, consistent with the dense and spatially pervasive distribution of diffuse parenchymal amyloid observed histologically.

```{r}
compute_cell_distances <- function(seurat_obj, caa_coords, paren_coords) {
  coords_df <- GetTissueCoordinates(seurat_obj) |>
    as.data.frame() |>
    dplyr::select(cell, x, y)

  cell_mat  <- coords_df |> dplyr::select(x, y) |> as.matrix()
  caa_mat   <- caa_coords   |> dplyr::select(x_plaque, y_plaque) |> as.matrix()
  paren_mat <- paren_coords |> dplyr::select(x_plaque, y_plaque) |> as.matrix()

  caa_nn   <- nn2(data = caa_mat,   query = cell_mat, k = 1)
  paren_nn <- nn2(data = paren_mat, query = cell_mat, k = 1)

  coords_df |>
    mutate(
      dist_caa         = caa_nn$nn.dist[, 1],
      nearest_caa_id   = caa_coords$plaque_id[caa_nn$nn.idx[, 1]],
      dist_paren       = paren_nn$nn.dist[, 1],
      nearest_paren_id = paren_coords$plaque_id[paren_nn$nn.idx[, 1]]
    )
}

cell_distances_1 <- compute_cell_distances(slide1, caa_coords_1, paren_coords_1)
cell_distances_2 <- compute_cell_distances(slide2, caa_coords_2, paren_coords_2)
```

```{r}
#| fig-height: 4
bind_rows(cell_distances_1, cell_distances_2) |>
  pivot_longer(c(dist_caa, dist_paren), names_to = "plaque_type", values_to = "min_dist") |>
  ggplot(aes(x = min_dist)) +
  geom_histogram(binwidth = 50, fill = "lightgray", col = "black") +
  labs(x = "min. dist (µm)") +
  facet_wrap(~plaque_type, scales = "free") +
  theme_minimal() +
  theme(strip.text = element_text(size = 14))
```

## Proximity classification

Each cell was assigned to either the CAA or parenchymal plaque neighbourhood *based on which plaque type was spatially closer*. Across the six samples (n = 51,959–69,664 cells per brain), the large majority of cells were classified as parenchymal-proximal (87.5–92.9% per sample), with 7.4–12.5% classified as CAA-proximal, broadly reflecting the disparity in plaque abundance between types. Between 4.1–6.6% of cells per sample fell within the 50 µm CAA proximity threshold, while 38.5–52.2% of cells were within 50 µm of a parenchymal plaque.



Each cell is assigned to whichever plaque type is nearest: **CAA** if `dist_caa < dist_paren`, **Parenchymal** otherwise.

```{r}
classify_proximity <- function(dist_data) {
  dist_data |>
    column_to_rownames("cell") |>
    mutate(
      caa = case_when(
        dist_caa < 50  ~ "<50um",
        dist_caa >= 50 ~ "≥50um"),
      parenchymal = case_when(
        dist_paren < 50                    ~ "<50um",
        between(dist_paren, 50, 150)       ~ "50-150um",
        dist_paren >= 150                  ~ "≥150um"),
      classification = if_else(dist_caa < dist_paren, "CAA", "Parenchymal")
    ) |>
    select(dist_caa:classification)
}

slide1 <- AddMetaData(slide1, classify_proximity(cell_distances_1))
slide2 <- AddMetaData(slide2, classify_proximity(cell_distances_2))
```

```{r}
md <- bind_rows(slide1@meta.data, slide2@meta.data)

c1 <- compareGroups(sample_ID ~ caa + parenchymal + classification, data = md, max.ylev = 6)
export2md(createTable(c1, show.p.overall = FALSE), caption = "")
```

At the treatment-group level, statistically significant differences in spatial proximity were detected between Aducanumab-treated (n=179,686 cells) and IgG-control (n=178,713 cells) animals. A higher proportion of IgG cells were in close proximity (<50 µm) to parenchymal plaques compared to Aducanumab-treated animals (48.0% vs. 42.5%; p<0.001), while Aducanumab-treated animals had a greater fraction of cells at intermediate-to-distal parenchymal distances (50–150 µm: 45.7% vs. 42.5%; >150 µm: 11.8% vs. 9.54%). The proportion of cells classified as CAA-proximal was modestly but significantly higher in the IgG group (9.53% vs. 9.01%; p<0.001). These findings suggest that Aducanumab treatment is associated with a subtle but statistically detectable reorganisation of the cellular microenvironment around amyloid deposits. Note that the large cell counts (n > 100k per group) confer high statistical power to detect even small differences; effect sizes should be interpreted accordingly, and downstream analyses using vascular gene expression scores will be essential to determine whether these spatial shifts are biologically meaningful.

```{r}
c2 <- compareGroups(Treatment.Group ~ caa + parenchymal + classification, data = md)
export2md(createTable(c2), caption = "")
```

_Note that the large cell counts (n > 100k per group) confer high statistical power to detect even small differences. Effect sizes should be interpreted accordingly._