---
title: "Co-registration — Akhil (single annotator)"
subtitle: "ThioS quantification and HALO analysis"
date: last-modified
execute:
  echo: true
  warning: false
embed-resources: true
format: html
code-fold: true
tags: [xenium, coregistration, halo]
---

```{r}
source("R/setup.R")
library(compareGroups)

data <- load_slides_and_distances("full")
slide1 <- data$slide1; slide2 <- data$slide2
cell_distances_1 <- data$distances_1; cell_distances_2 <- data$distances_2
```

# HALO analysis

All annotations performed by Akhil Palleria (single annotator; no inter-observer variability).

## Metrics per sample

Thioflavin-S (ThioS)-positive amyloid plaques were quantified across six brain samples (Brain1–6; IgG control, n=3; Aducanumab-treated, n=3) using HALO image analysis software, with all annotations performed by a single rater (A.P.) to eliminate inter-observer variability.  

Two plaque phenotypes were identified: cerebrovascular (CAA) and parenchymal plaques. Across all brains, parenchymal plaques were substantially more numerous than CAA plaques, with per-brain counts ranging from 2,681 to 3,961 (parenchymal) versus 284 to 446 (CAA). Consistent with this, parenchymal plaques accounted for the majority of total annotated area in every sample.



```{r}
t1 <- bind_rows(thios_1, thios_2) |>
  group_by(brain, sample, plaque_type) |>
  summarise(
    n_plaques  = n(),
    total_area = sum(Area),
    .groups    = "drop"
  )

t1 |> kbl() |> kable_styling("striped")
```

Group-level comparison revealed no statistically significant difference between Aducanumab-treated and IgG-control animals in either total CAA area (41,759 ± 17,904 vs. 39,697 ± 6,076 µm²; p=0.865), total parenchymal plaque area (410,021 ± 121,107 vs. 374,302 ± 40,078 µm²; p=0.668), or the proportion of total plaque burden attributable to CAA (8.97 ± 1.32% vs. 9.69 ± 2.13%; p=0.653).
```{r}
t1_wide <- t1 |>
  pivot_wider(id_cols = c(brain, sample), names_from = plaque_type, values_from = total_area) |>
  mutate(CAA_percent = CAA / (Parenchymal + CAA) * 100) |>
  inner_join(sample_dict, by = c("brain", "sample"))

caa1 <- compareGroups(treatment ~ CAA + Parenchymal + CAA_percent, data = t1_wide)
export2md(createTable(caa1, show.p.overall = T), caption = "")

```

## Plaque centroids

To enable spatial co-registration with single-cell Xenium transcriptomic data, the centroid of each plaque was computed as the midpoint of the HALO bounding box and converted from pixel coordinates to micrometres using a calibration factor of 0.2125 µm/pixel. This coordinate transformation provides a common spatial reference frame for subsequent nearest-neighbour analyses.

Centroid and distance functions are defined in `R/setup.R` and computed by `load_slides_and_distances()`.

## Cell-to-plaque distances (k-d tree)

Minimum Euclidean distances from **each Xenium-profiled** cell to its nearest CAA plaque (dist_caa) and nearest parenchymal plaque (dist_paren) were computed using a k-d tree algorithm. The resulting distance distributions were markedly different between plaque types. For CAA plaques, the distribution had a modal distance of approximately 200–300 µm and a long tail extending to ~1,500 µm, reflecting the relatively sparse and focal distribution of vascular amyloid deposits. By contrast, the majority of cells were within ~50 µm of a parenchymal plaque, consistent with the dense and spatially pervasive distribution of diffuse parenchymal amyloid observed histologically.

```{r}
#| fig-height: 4
bind_rows(cell_distances_1, cell_distances_2) |>
  pivot_longer(c(dist_caa, dist_paren), names_to = "plaque_type", values_to = "min_dist") |>
  ggplot(aes(x = min_dist)) +
  geom_histogram(binwidth = 50, fill = "lightgray", col = "black") +
  labs(x = "min. dist (µm)") +
  facet_wrap(~plaque_type, scales = "free") +
  theme_minimal() +
  theme(strip.text = element_text(size = 14))
```

## Proximity classification

Each cell was assigned to either the CAA or parenchymal plaque neighbourhood *based on which plaque type was spatially closer*. Across the six samples (n = 51,959–69,664 cells per brain), the large majority of cells were classified as parenchymal-proximal (87.5–92.9% per sample), with 7.4–12.5% classified as CAA-proximal, broadly reflecting the disparity in plaque abundance between types. Between 4.1–6.6% of cells per sample fell within the 50 µm CAA proximity threshold, while 38.5–52.2% of cells were within 50 µm of a parenchymal plaque.



Each cell is assigned to whichever plaque type is nearest: **CAA** if `dist_caa < dist_paren`, **Parenchymal** otherwise.

```{r}
classify_proximity <- function(dist_data) {
  dist_data |>
    column_to_rownames("cell") |>
    mutate(
      caa = case_when(
        dist_caa < 50  ~ "<50um",
        dist_caa >= 50 ~ "≥50um"),
      parenchymal = case_when(
        dist_paren < 50                        ~ "<50um",
        dist_paren >= 50 & dist_paren < 150   ~ "50-150um",
        dist_paren >= 150                      ~ "≥150um"),
      classification = if_else(dist_caa < dist_paren, "CAA", "Parenchymal")
    ) |>
    select(dist_caa:classification)
}

slide1 <- AddMetaData(slide1, classify_proximity(cell_distances_1))
slide2 <- AddMetaData(slide2, classify_proximity(cell_distances_2))
```

```{r}
md <- bind_rows(slide1@meta.data, slide2@meta.data)

c1 <- compareGroups(sample_ID ~ caa + parenchymal + classification, data = md, max.ylev = 6)
export2md(createTable(c1, show.p.overall = FALSE), caption = "")
```

**Caveat — pseudoreplication.** The cell-level p-values in the table below treat each cell as an independent observation, but cells from the same brain are not independent. The true experimental unit is the brain (n = 3 Adu vs 3 IgG), so these p-values are inflated by pseudoreplication and should be treated as descriptive summaries only. Formal treatment comparisons use pseudobulk aggregation (see notebooks 7 and 8). Consult the computed table for exact numbers.

```{r}
c2 <- compareGroups(Treatment.Group ~ caa + parenchymal + classification, data = md)
export2md(createTable(c2), caption = "")
```

_**Pseudoreplication warning:** these cell-level statistics treat each cell as independent, but the true unit of replication is the brain (n = 3 per group). P-values are therefore inflated and should not be used for formal inference. See pseudobulk DE in notebooks 7 and 8 for properly replicated treatment comparisons._