---
title: "Gene-to-distance-to-plaque correlation"
subtitle: "Replication of Saito et al. (2025) approach w/ Akhil annotations"
date: last-modified
execute:
  echo: true
  warning: false
embed-resources: true
format: html
code-fold: true
tags: [xenium, coregistration, halo, correlation]
---

# Setup and data loading

Load libraries, Seurat objects, and compute cell-to-plaque distances.

```{r}
#| label: setup
source("R/setup.R")
library(ggrepel)

data <- load_slides_and_distances("full")
slide1 <- data$slide1; slide2 <- data$slide2
cell_distances_1 <- data$distances_1; cell_distances_2 <- data$distances_2
cell_distances_all <- data$distances_all
```

# Gene-distance correlation functions

**does a cell's expression of this gene change systematically as the cell sits closer to (or farther from) a plaque?**

## `compute_gene_distance_correlation()`


For each gene we compute a **Spearman rank correlation (rho)** between two vectors that share one entry per cell:

1. The cell's raw transcript count for that gene.
2. The cell's Euclidean distance (in um) to its nearest plaque of the specified type.

**Spearman correlation** operates on ranks rather than raw values, so it captures any monotonic relationship (not just linear) and is robust to the extreme zero-inflation typical of spatial transcriptomics count data. We set `exact = FALSE` because the large number of tied zeros makes an exact permutation p-value computationally infeasible.

A gene that is highly expressed *near* plaques (short distance) and lowly expressed *far away* (long distance) will have a **negative** rho (high expression co-occurs with low distance). Conversely, a gene expressed preferentially *far from* plaques will have a **positive** rho.

To make the results more intuitive,so that "plaque-enriched" genes appear at the *top* of a ranked list rather than the bottom, we multiply rho by -1 to obtain `rho_flip`. After flipping, positive values mean "enriched near plaques" and negative values mean "enriched far from plaques".

 We apply the Benjamini-Hochberg (BH) procedure to control the false discovery rate, producing an adjusted p-value (`padj`). A gene is called significant at `padj < 0.05`.

**Final classification.** Each gene is labelled as:

- **enriched_near** — `padj < 0.05` and `rho_flip > 0` (upregulated close to plaques)
- **enriched_far** — `padj < 0.05` and `rho_flip < 0` (upregulated far from plaques)
- **ns** — not significant


```{r}
#| label: define-compute-correlation
compute_gene_distance_correlation <- function(
    seurat_list,
    cell_distances,
    distance_col = "dist_any",
    min_expressing_cells = 100
) {
  # ── 1. Pool raw count matrices across slides ────────────────────────────────
  # Use the RNA assay counts layer (Seurat v5 sparse matrices)
  count_list <- map(seurat_list, \(obj) {
    LayerData(obj, assay = DefaultAssay(obj), layer = "counts")
  })

  # Ensure same gene set, then column-bind
  shared_genes <- Reduce(intersect, map(count_list, rownames))
  counts <- do.call(cbind, map(count_list, \(m) m[shared_genes, ]))

  # ── 2. Align cells with distance table ──────────────────────────────────────
  shared_cells <- intersect(colnames(counts), cell_distances$cell)
  counts <- counts[, shared_cells]
  dist_vec <- cell_distances |>
    filter(cell %in% shared_cells) |>
    arrange(match(cell, shared_cells)) |>
    pull(!!sym(distance_col))

  # ── 3. Filter low-expressed genes ───────────────────────────────────────────
  n_expressing <- Matrix::rowSums(counts > 0)
  keep <- n_expressing >= min_expressing_cells
  counts <- counts[keep, ]
  message(
    sprintf(
      "Kept %d / %d genes (≥%d expressing cells)",
      sum(keep), length(keep), min_expressing_cells
    )
  )

  # ── 4. Per-gene Spearman correlation ────────────────────────────────────────
  # exact = FALSE to handle tied ranks from sparse zeros efficiently
  results <- map(seq_len(nrow(counts)), \(i) {
    gene_expr <- counts[i, ]
    ct <- cor.test(as.numeric(gene_expr), dist_vec,
                   method = "spearman", exact = FALSE)
    tibble(
      gene = rownames(counts)[i],
      rho  = ct$estimate,
      pval = ct$p.value
    )
  }, .progress = "Spearman correlations") |>
    bind_rows()

  # ── 5. BH-correct, flip sign, rank ─────────────────────────────────────────
  # Flip rho so that genes enriched near plaques (negative rho) become positive
  results |>
    mutate(
      padj      = p.adjust(pval, method = "BH"),
      rho_flip  = -rho,
      rank      = rank(-rho_flip, ties.method = "first"),
      sig       = case_when(
        padj < 0.05 & rho_flip > 0 ~ "enriched_near",
        padj < 0.05 & rho_flip < 0 ~ "enriched_far",
        TRUE                        ~ "ns"
      )
    ) |>
    arrange(rank)
}
```

## `plot_gene_distance_correlation()`

Generates a rank plot. Each dot is one gene.  

The x-axis is the gene's rank (1 = strongest plaque enrichment) and the y-axis is `rho_flip` (the sign-flipped Spearman correlation). Genes sitting high on the plot are preferentially expressed near plaques; genes at the bottom are expressed far from plaques. The dashed line at zero marks no association.   

Orange = significantly enriched near plaques; purple = significantly enriched far from plaques; grey = not significant (BH-adjusted p >= 0.05). The top and bottom genes are automatically labelled, plus any user-specified genes of interest.

```{r}
#| label: define-plot-correlation
plot_gene_distance_correlation <- function(
    results,
    title         = "Gene–distance correlation",
    n_top         = 5,
    genes_to_label = NULL
) {
  # Auto-label top n positive and negative genes
  top_pos <- results |> filter(sig == "enriched_near") |> slice_min(rank, n = n_top)
  top_neg <- results |> filter(sig == "enriched_far")  |> slice_max(rank, n = n_top)
  auto_label <- bind_rows(top_pos, top_neg)

  # Add user-specified genes
  if (!is.null(genes_to_label)) {
    user_genes <- results |> filter(gene %in% genes_to_label)
    auto_label <- bind_rows(auto_label, user_genes) |> distinct(gene, .keep_all = TRUE)
  }

  # Colour mapping
  colour_map <- c(
    "enriched_near" = "#E07B39",
    "enriched_far"  = "#7B68AE",
    "ns"            = "grey70"
  )

  ggplot(results, aes(x = rank, y = rho_flip)) +
    geom_point(aes(colour = sig), size = 2, alpha = 0.5) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
    geom_text_repel(
      data          = auto_label,
      aes(label = gene, colour = sig),
      size          = 4,
      fontface      = "italic",
      label.size    = 0.2,
      box.padding   = 0.4,
      max.overlaps  = 20,
      show.legend   = FALSE
    ) +
    scale_colour_manual(
      values = colour_map,
      labels = c(
        "enriched_near" = "Enriched near plaque (padj < 0.05)",
        "enriched_far"  = "Enriched far from plaque (padj < 0.05)",
        "ns"            = "Not significant"
      )
    ) +
    labs(
      title  = title,
      x      = "Gene rank",
      y      = "Spearman rho (flipped)",
      colour = NULL
    ) +
    theme_minimal(base_size = 14) +
    theme(
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank())
}
```

# Run analysis

Compute gene–distance correlations for three distance metrics:

- **dist_caa** — distance to nearest CAA plaque
- **dist_paren** — distance to nearest parenchymal plaque
- **dist_any** — distance to nearest plaque of any type

```{r}
#| label: run-correlations
seurat_list <- list(slide1, slide2)

cor_caa <- compute_gene_distance_correlation(
  seurat_list, cell_distances_all, distance_col = "dist_caa"
)

cor_paren <- compute_gene_distance_correlation(
  seurat_list, cell_distances_all, distance_col = "dist_paren"
)

cor_any <- compute_gene_distance_correlation(
  seurat_list, cell_distances_all, distance_col = "dist_any"
)
```

## Rank plots


Three analyses are shown, each using a different distance metric: (1) distance to nearest **CAA** plaque only, (2) distance to nearest **parenchymal** plaque only, and (3) distance to nearest plaque of **any type**. Comparing across the three reveals whether certain genes are specifically associated with one plaque subtype or with amyloid deposits in general.

```{r}
#| label: plot-caa
#| fig-height: 5
#| fig-width: 6
#| fig-cap: "For each gene, Spearman’s rank correlation coefficient (ρ) was calculated between per-cell gene expression levels (raw count data) and the minimum distance to the nearest amyloid plaque. To facilitate interpretation, correlation coefficients were sign-inverted such that positive values indicate higher gene expression in cells located closer to plaques, whereas negative values indicate higher expression at increasing distances from plaques. Genes were ranked by the inverted correlation coefficient. Statistical significance was assessed using two-sided Spearman correlation tests followed by Benjamini–Hochberg false discovery rate correction. Genes were classified as plaque-enriched, plaque-distal, or not significant based on adjusted P values. The five most positive and five most negative correlations are labeled"
plot_gene_distance_correlation(cor_caa, title = "Distance to nearest CAA plaque")
```

```{r}
#| label: plot-paren
#| fig-height: 5
#| fig-width: 6
plot_gene_distance_correlation(cor_paren, title = "Distance to nearest parenchymal plaque")
```

```{r}
#| label: plot-any
#| fig-height: 5
#| fig-width: 6
plot_gene_distance_correlation(cor_any, title = "Distance to nearest plaque (any type)")
```

## Top genes

```{r}
#| label: tables
show_top <- function(results, label) {
  top <- bind_rows(
    results |> filter(sig == "enriched_near") |> slice_min(rank, n = 10),
    results |> filter(sig == "enriched_far")  |> slice_max(rank, n = 10)
  ) |>
    select(gene, rho, rho_flip, pval, padj, rank, sig) |>
    mutate(across(c(rho, rho_flip), \(x) round(x, 4)),
           across(c(pval, padj), \(x) ifelse(x < 1e-300, "< 1e-300", formatC(x, format = "e", digits = 2))))

  top |>
    kbl(caption = label) |>
    kable_styling("striped", full_width = FALSE)
}

show_top(cor_caa,   "Top genes — distance to CAA")
show_top(cor_paren, "Top genes — distance to parenchymal")
show_top(cor_any,   "Top genes — distance to any plaque")
```

