---
title: "Gene-to-distance-to-plaque correlation — microglia only"
subtitle: "Microglia object provided by Kai Saito with  Akhil HALO annotations"
date: last-modified
---


By restricting the analysis to **microglia only**, we remove the composition
confound. If a gene still correlates with distance within a single cell type,
the association more likely reflects a genuine distance-dependent
transcriptional programme rather than the fact that microglia are (or are not)
concentrated near plaques.

# Setup and data loading

```{r}
#| label: setup
source("R/setup.R")
library(ggrepel)
library(patchwork)
library(ggVennDiagram)
library(gridExtra)

data <- load_slides_and_distances("microglia")
slide1_mg <- data$slide1; slide2_mg <- data$slide2
cell_distances_1 <- data$distances_1; cell_distances_2 <- data$distances_2
cell_distances_all <- data$distances_all

message(sprintf(
  "Microglia: slide 1 = %d cells, slide 2 = %d cells (total = %d)",
  ncol(slide1_mg), ncol(slide2_mg), ncol(slide1_mg) + ncol(slide2_mg)
))
```

```{r}
#| label: add-proximity-metadata
# Add proximity labels (50 um threshold) — needed for treatment DE later
add_proximity <- function(seurat_obj, cell_distances) {
  prox <- cell_distances |>
    mutate(
      prox_caa   = if_else(dist_caa < 50,   "proximal", "distal"),
      prox_paren = if_else(dist_paren < 50,  "proximal", "distal"),
      prox_any   = if_else(pmin(dist_caa, dist_paren) < 50, "proximal", "distal")
    ) |>
    column_to_rownames("cell") |>
    select(dist_caa, dist_paren, prox_caa, prox_paren, prox_any)

  AddMetaData(seurat_obj, prox)
}

slide1_mg <- add_proximity(slide1_mg, cell_distances_1)
slide2_mg <- add_proximity(slide2_mg, cell_distances_2)

cell_distances_all <- cell_distances_all |>
  mutate(
    prox_caa   = if_else(dist_caa < 50,   "proximal", "distal"),
    prox_paren = if_else(dist_paren < 50,  "proximal", "distal"),
    prox_any   = if_else(dist_any < 50,    "proximal", "distal")
  )
```

## Cell counts per group

```{r}
#| label: cell-counts
# sample_ID and Treatment.Group are already in cell_distances_all (joined in load_slides_and_distances)
cell_distances_all |>
  group_by(sample_ID, Treatment.Group, prox_any) |>
  summarise(n = n(), .groups = "drop") |>
  pivot_wider(names_from = prox_any, values_from = n, values_fill = 0) |>
  mutate(total = proximal + distal) |>
  kbl(caption = "Microglia per sample — proximity to any plaque (50 um threshold)") |>
  kable_styling("striped", full_width = FALSE)
```


# Correlation function and plot

For each gene we compute a **Spearman rank correlation (rho)** between two vectors that share one entry per cell:

1. The cell's raw transcript count for that gene.
2. The cell's Euclidean distance (in um) to its nearest plaque of the specified type.

**Spearman correlation** operates on ranks rather than raw values, so it captures any monotonic relationship (not just linear) and is robust to the extreme zero-inflation typical of spatial transcriptomics count data. We set `exact = FALSE` because the large number of tied zeros makes an exact permutation p-value computationally infeasible.

A gene that is highly expressed *near* plaques (short distance) and lowly expressed *far away* (long distance) will have a **negative** rho (high expression co-occurs with low distance). Conversely, a gene expressed preferentially *far from* plaques will have a **positive** rho.

To make the results more intuitive,so that "plaque-enriched" genes appear at the *top* of a ranked list rather than the bottom, we multiply rho by -1 to obtain `rho_flip`. After flipping, positive values mean "enriched near plaques" and negative values mean "enriched far from plaques".

 We apply the Benjamini-Hochberg (BH) procedure to control the false discovery rate, producing an adjusted p-value (`padj`). A gene is called significant at `padj < 0.05`.

**Final classification.** Each gene is labelled as:

- **enriched_near** — `padj < 0.05` and `rho_flip > 0` (upregulated close to plaques)
- **enriched_far** — `padj < 0.05` and `rho_flip < 0` (upregulated far from plaques)
- **ns** — not significant

`compute_gene_distance_correlation()` is centralised in `R/setup.R` (sourced in setup chunk).

```{r}
#| label: define-plot-correlation
plot_gene_distance_correlation <- function(
    results,
    title          = "Gene-distance correlation",
    n_top          = 5,
    genes_to_label = NULL
) {
  top_pos <- results |> filter(sig == "enriched_near") |> slice_min(rank, n = n_top)
  top_neg <- results |> filter(sig == "enriched_far")  |> slice_max(rank, n = n_top)
  auto_label <- bind_rows(top_pos, top_neg)

  if (!is.null(genes_to_label)) {
    user_genes <- results |> filter(gene %in% genes_to_label)
    auto_label <- bind_rows(auto_label, user_genes) |> distinct(gene, .keep_all = TRUE)
  }

  colour_map <- c(
    "enriched_near" = "#E07B39",
    "enriched_far"  = "#7B68AE",
    "ns"            = "grey70"
  )

  ggplot(results, aes(x = rank, y = rho_flip)) +
    geom_point(aes(colour = sig), size = 2, alpha = 0.5) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
    geom_text_repel(
      data          = auto_label,
      aes(label = gene, colour = sig),
      size          = 5,
      fontface      = "italic",
      box.padding   = 0.4,
      max.overlaps  = 20,
      show.legend   = FALSE,
      seed = 23
    ) +
    scale_colour_manual(
      values = colour_map,
      labels = c(
        "enriched_near" = "Enriched near plaque (padj < 0.05)",
        "enriched_far"  = "Enriched far from plaque (padj < 0.05)",
        "ns"            = "Not significant"
      )
    ) +
    labs(
      title  = title,
      x      = "Gene rank",
      y      = "Spearman rho (flipped)",
      colour = NULL
    ) +
    theme_minimal(base_size = 14) +
    theme(
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank())
}
```

# Run analysis — microglia only

```{r}
#| label: run-correlations
#| eval: false
seurat_list <- list(slide1_mg, slide2_mg)

cor_caa   <- compute_gene_distance_correlation(seurat_list, cell_distances_all, distance_col = "dist_caa")
cor_paren <- compute_gene_distance_correlation(seurat_list, cell_distances_all, distance_col = "dist_paren")
cor_any   <- compute_gene_distance_correlation(seurat_list, cell_distances_all, distance_col = "dist_any")

# qs2::qs_save(cor_caa,   "spatial/mg_cor_caa.qs2")
# qs2::qs_save(cor_paren, "spatial/mg_cor_paren.qs2")
# qs2::qs_save(cor_any,   "spatial/mg_cor_any.qs2")
```

```{r}
#| label: run-correlations-by-treatment
#| eval: false
seurat_list <- list(slide1_mg, slide2_mg)

dist_adu <- cell_distances_all |> filter(Treatment.Group == "Adu")
dist_igg <- cell_distances_all |> filter(Treatment.Group == "IgG")

for (trt in c("adu", "igg")) {
  dist_trt <- if (trt == "adu") dist_adu else dist_igg
  for (dcol in c("dist_caa", "dist_paren", "dist_any")) {
    res    <- compute_gene_distance_correlation(seurat_list, dist_trt, distance_col = dcol)
    suffix <- sub("dist_", "", dcol)
    qs2::qs_save(res, paste0("spatial/mg_cor_", suffix, "_", trt, ".qs2"))
  }
}
```

```{r}
#| label: load-correlations

cor_caa   <- qs2::qs_read("spatial/mg_cor_caa.qs2")
cor_paren <- qs2::qs_read("spatial/mg_cor_paren.qs2")
cor_any   <- qs2::qs_read("spatial/mg_cor_any.qs2")
```

## Rank plots

The microglia-specific gene-distance correlation, split by plaque type reveals that microglia activate fundamentally different transcriptional programmes near CAA vs parenchymal plaques.      
                                                                                                                                                                                                            
**Parenchymal plaque (right panel)**                                                                                                                                                                          
                                                                                                                                                                                                                                                                                                                                                                                                         
Near plaque (orange): Classic **DAM** (Disease-Associated Microglia) signature:

  - Cst7 (rho ~0.45) — the canonical DAM marker, cystatin F, involved in lysosomal processing
  - Ctsb, Cd63 — lysosomal/endosomal genes: microglia near parenchymal plaques are actively phagocytosing amyloid
  - Lyz2 — lysozyme, antimicrobial/inflammatory
  - Lpl — lipoprotein lipase, lipid metabolism rewiring in DAM

  Far from plaque (purple): **Homeostatic** microglia signature:

  - P2ry12, Tmem119 — the two defining homeostatic microglia markers. Their strong negative rho confirms that microglia lose their resting identity as they approach parenchymal plaques.
  - Maf, Ptgs1, Lpcat2 — transcription factor and lipid metabolism genes associated with the surveilling/quiescent state

 

**CAA plaque (left panel)**

_The effect sizes are much smaller (~0.08 vs ~0.45) and the gene signature is completely different_.

Near CAA (orange): This is a lipid/cholesterol metabolism programme, not a phagocytic one. CAA sits in vessel walls, and microglia nearby appear to be responding to vascular lipid dysregulation rather than engulfing
  amyloid.

  - Glul — glutamine synthetase, typically an astrocyte gene but also involved in glutamate-glutamine cycling at the vasculature
  - Abhd17a — depalmitoylase, involved in protein trafficking
  - Srebf2, Hmgcs1 — cholesterol biosynthesis pathway (SREBP2 is the master regulator, HMGCS1 is its direct target)
  - Acsl6 — fatty acid activation

  
  Far from CAA (purple): 

  - Lgals3bp, Ftl1 — galectin-binding and ferritin (iron storage)
  - Abca1, Abcg1 — cholesterol efflux transporters (ABC family). 


```{r}
#| label: plot-caa-paren-combined
#| column: page
#| fig-height: 5
#| fig-width: 10
#| fig-cap: "For each gene, Spearman’s rank correlation coefficient (ρ) was calculated between per-cell gene expression levels (raw count data) and the minimum distance to the nearest amyloid plaque. To facilitate interpretation, correlation coefficients were sign-inverted such that positive values indicate higher gene expression in cells located closer to plaques, whereas negative values indicate higher expression at increasing distances from plaques. Genes were ranked by the inverted correlation coefficient. Statistical significance was assessed using two-sided Spearman correlation tests followed by Benjamini–Hochberg false discovery rate correction. Genes were classified as plaque-enriched, plaque-distal, or not significant based on adjusted P values. The five most positive and five most negative correlations are labeled"

p_caa   <- plot_gene_distance_correlation(cor_caa,   title = "CAA plaque")
p_paren <- plot_gene_distance_correlation(cor_paren, title = "Parenchymal plaque")

(p_caa | p_paren) +
  plot_annotation( title = "Plaque proximity gene-distance correlation")+
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

```{r}
#| label: venn-enriched-near
#| fig-height: 4
#| fig-width: 6
#| fig-cap: "Overlap of genes significantly enriched near plaques (padj < 0.05, positive rho_flip) between CAA and parenchymal analyses."

near_caa   <- cor_caa   |> filter(padj < 0.05, rho_flip > 0) |> pull(gene)
near_paren <- cor_paren |> filter(padj < 0.05, rho_flip > 0) |> pull(gene)

ggVennDiagram(
  list(CAA = near_caa, Parenchymal = near_paren),
  label_alpha = 0, label = "count", shape_id = "201") +
  labs(title = "Enriched-near genes (padj < 0.05)") +
  theme(legend.position = "none")+
  scale_fill_distiller(palette = "RdBu")

cat("Common enriched-near genes in both CAA and parenchymal analyses:\n",
  paste(intersect(near_caa, near_paren), collapse = ", "))

```

```{r}
#| label: plot-any
#| fig-height: 5
#| fig-width: 6
plot_gene_distance_correlation(cor_any, title = "Microglia — distance to nearest plaque (any type)")
```

## Top genes

```{r}
#| label: tables
show_top <- function(results, label) {
  top <- bind_rows(
    results |> filter(sig == "enriched_near") |> slice_min(rank, n = 10),
    results |> filter(sig == "enriched_far")  |> slice_max(rank, n = 10)
  ) |>
    select(gene, rho, rho_flip, pval, padj, rank, sig) |>
    mutate(across(c(rho, rho_flip), \(x) round(x, 4)),
           across(c(pval, padj), \(x) ifelse(x < 1e-300, "< 1e-300", formatC(x, format = "e", digits = 2))))

  top |>
    kbl(caption = label) |>
    kable_styling("striped", full_width = FALSE)
}

show_top(cor_caa,   "Top genes — microglia, distance to CAA")
show_top(cor_paren, "Top genes — microglia, distance to parenchymal")
show_top(cor_any,   "Top genes — microglia, distance to any plaque")
```

# Treatment-split correlations

Same analysis run separately in **Aducanumab** and **IgG** microglia.

```{r}
#| label: load-correlations-by-treatment

cor_caa_adu   <- qs2::qs_read("spatial/mg_cor_caa_adu.qs2")
cor_caa_igg   <- qs2::qs_read("spatial/mg_cor_caa_igg.qs2")
cor_paren_adu <- qs2::qs_read("spatial/mg_cor_paren_adu.qs2")
cor_paren_igg <- qs2::qs_read("spatial/mg_cor_paren_igg.qs2")
cor_any_adu   <- qs2::qs_read("spatial/mg_cor_any_adu.qs2")
cor_any_igg   <- qs2::qs_read("spatial/mg_cor_any_igg.qs2")
```

## Rank plots by treatment

```{r}
#| label: plot-caa-treatment
#| column: page
#| fig-height: 5
#| fig-width: 10
plot_gene_distance_correlation(cor_caa_adu, title = "CAA — Aducanumab") |
plot_gene_distance_correlation(cor_caa_igg, title = "CAA — IgG")
```

```{r}
#| label: plot-paren-treatment
#| column: page
#| fig-height: 5
#| fig-width: 10
plot_gene_distance_correlation(cor_paren_adu, title = "Parenchymal — Aducanumab") |
plot_gene_distance_correlation(cor_paren_igg, title = "Parenchymal — IgG")
```

```{r}
#| label: plot-any-treatment
#| column: page
#| fig-height: 5
#| fig-width: 10
plot_gene_distance_correlation(cor_any_adu, title = "Any plaque — Aducanumab") |
plot_gene_distance_correlation(cor_any_igg, title = "Any plaque — IgG")
```

## Adu vs IgG scatter

```{r}
#| label: plot-scatter-treatment
#| column: page
#| fig-height: 5
#| fig-width: 12

plot_treatment_scatter <- function(cor_adu, cor_igg, title) {
  df <- inner_join(
    cor_adu |> select(gene, rho_flip_adu = rho_flip, sig_adu = sig),
    cor_igg |> select(gene, rho_flip_igg = rho_flip, sig_igg = sig),
    by = "gene"
  ) |>
    mutate(
      concordance = case_when(
        sig_adu == "enriched_near" & sig_igg == "enriched_near" ~ "both near",
        sig_adu == "enriched_far"  & sig_igg == "enriched_far"  ~ "both far",
        sig_adu != "ns"            & sig_igg != "ns"            ~ "discordant",
        sig_adu != "ns"            | sig_igg != "ns"            ~ "one sig",
        TRUE                                                     ~ "ns"
      )
    )

  colour_map <- c(
    "both near"  = "#E07B39",
    "both far"   = "#7B68AE",
    "discordant" = "#C0392B",
    "one sig"    = "#5DADE2",
    "ns"         = "grey80"
  )

  label_df <- df |> filter(concordance %in% c("both near", "both far", "discordant"))

  ggplot(df, aes(x = rho_flip_igg, y = rho_flip_adu)) +
    geom_point(aes(colour = concordance), size = 1.8, alpha = 0.6) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
    geom_hline(yintercept = 0, colour = "grey70", linewidth = 0.3) +
    geom_vline(xintercept = 0, colour = "grey70", linewidth = 0.3) +
    geom_text_repel(
      data         = label_df,
      aes(label = gene, colour = concordance),
      size         = 3.5,
      fontface     = "italic",
      max.overlaps = 20,
      show.legend  = FALSE
    ) +
    scale_colour_manual(values = colour_map) +
    labs(
      title  = title,
      x      = "rho_flip — IgG",
      y      = "rho_flip — Aducanumab",
      colour = NULL
    ) +
    theme_minimal(base_size = 13) +
    theme(
      legend.position  = "bottom",
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "grey92")
    )
}

plot_treatment_scatter(cor_caa_adu,   cor_caa_igg,   "CAA — Adu vs IgG") |
plot_treatment_scatter(cor_paren_adu, cor_paren_igg, "Parenchymal — Adu vs IgG") |
plot_treatment_scatter(cor_any_adu,   cor_any_igg,   "Any plaque — Adu vs IgG")
```

# Candidate gene heatmap

The rank plots above identified two biologically distinct distance-dependent programmes. To compare their effect sizes and significance side by side across both plaque types, we visualise the rho_flip values for the named genes in a single heatmap. Candidates were drawn directly from the labelled genes in the rank plots:

- **DAM / phagocytic programme** (near parenchymal plaque): Cst7, Ctsb, Cd63, Lyz2, Lpl — the top-ranked enriched-near genes for parenchymal distance.
- **Vascular-lipid programme** (near CAA): Srebf2, Hmgcs1, Acsl6, Glul, Abhd17a — the top-ranked enriched-near genes for CAA distance.
- **Homeostatic programme** (far from parenchymal plaque): P2ry12, Tmem119, Maf, Ptgs1, Lpcat2 — the top-ranked enriched-far genes for parenchymal distance.
- **Cholesterol efflux / iron** (far from CAA): Abca1, Abcg1, Lgals3bp, Ftl1 — the top-ranked enriched-far genes for CAA distance.

The heatmap is intended to show whether genes that are strongly associated with one plaque type also show a weaker or opposing signal for the other.

```{r}
#| label: candidate-gene-heatmap
#| fig-height: 7
#| fig-width: 6
#| fig-cap: "Spearman rho_flip for microglia candidate genes across CAA and parenchymal distance. Positive (orange) = enriched near plaque; negative (purple) = enriched far from plaque. Stars indicate significance (* padj < 0.05, ** padj < 0.01, *** padj < 0.001)."

mg_candidate_genes <- c(
  "Cst7", "Ctsb", "Cd63", "Lyz2", "Lpl",
  "Srebf2", "Hmgcs1", "Acsl6", "Glul", "Abhd17a",
  "P2ry12", "Tmem119", "Maf", "Ptgs1", "Lpcat2",
  "Abca1", "Abcg1", "Lgals3bp", "Ftl1"
)

heatmap_df <- bind_rows(
  cor_caa   |> mutate(plaque = "CAA"),
  cor_paren |> mutate(plaque = "Parenchymal")
) |>
  filter(gene %in% mg_candidate_genes) |>
  mutate(
    sig_label = case_when(
      padj < 0.001 ~ "***",
      padj < 0.01  ~ "**",
      padj < 0.05  ~ "*",
      TRUE         ~ ""
    ),
    gene = factor(gene, levels = rev(mg_candidate_genes))
  )

ggplot(heatmap_df, aes(x = plaque, y = gene, fill = rho_flip)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = sig_label), size = 4, vjust = 0.75) +
  scale_fill_gradient2(
    low  = "#7B68AE", mid = "white", high = "#E07B39",
    midpoint = 0,
    name = "rho_flip\n(+ = near plaque)"
  ) +
  labs(
    title    = "Microglia candidate gene-distance correlations",
    subtitle = "* padj < 0.05  ** padj < 0.01  *** padj < 0.001",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

