---
title: "Microglia proximal vs distal to plaques "
subtitle: "Differential expression (pseudobulk DE) within microglia stratified by plaque proximity"
date: last-modified
execute:
  echo: true
  warning: false
embed-resources: true
format: html
code-fold: true
tags: [xenium, coregistration, halo, microglia, DE]
---

# Rationale

Instead of treating distance as a continuous variable,
we **dichotomize** microglia into proximal (< 50 um from a plaque) and distal
(>= 50 um) groups, then run **pseudobulk differential expression** (DESeq2)
comparing the two.

By aggregating counts per biological replicate (sample/brain) before testing, we
respect the experimental design (pseudobulk). Each mouse is the true unit of replication
(n = 6 brains, 3 per treatment).


# Setup and data loading

```{r}
#| label: setup
library(readxl)
library(tidyverse)
library(kableExtra)
library(Seurat)
library(RANN)
library(DESeq2)
library(EnhancedVolcano)
library(patchwork)
library(gridExtra)
library(ggridges)
options(scipen = 999)
```

```{r}
#| label: load-halo
# Read CAA plaque IDs from all sheets (one per brain) in Akhil's Excel file
caa_path <- "HALO_thioS_Akhil/20260217_CAAquant_AP.xlsx"

caa_akhil <- excel_sheets(caa_path) |>
  set_names() |>
  map(\(sheet) read_excel(caa_path, sheet = sheet)) |>
  bind_rows(.id = "sheet") |>
  mutate(
    brain = str_remove(sheet, "_AP"),
    index = paste0(tolower(brain), "_", Id)
  ) |>
  pull(index)

# Brain-to-sample mapping (treatment and slide assignment)
sample_dict <- tribble(
  ~brain,   ~sample,   ~treatment, ~slide,
  "Brain1", "KK4_465", "IgG",      1L,
  "Brain2", "KK4_492", "Adu",      1L,
  "Brain3", "KK4_504", "IgG",      1L,
  "brain4", "KK4_496", "IgG",      2L,
  "brain5", "KK4_502", "Adu",      2L,
  "brain6", "KK4_464", "Adu",      2L
)

# Read HALO plaque segmentation CSVs, classify each plaque as CAA or Parenchymal
thios_all <- list.files(
  "HALO_thioS_Akhil",
  pattern    = "^20260220.*AP\\.csv$",
  full.names = TRUE
) |>
  map(\(x) read_csv(x, locale = locale(encoding = "latin1"))) |>
  bind_rows() |>
  rename_with(\(x) str_remove(x, "\\s*\\(.*\\)$")) |>
  dplyr::rename(brain = "Analysis Region", objectID = "Object Id") |>
  mutate(
    index       = paste0(tolower(brain), "_", objectID),
    plaque_type = if_else(index %in% caa_akhil, "CAA", "Parenchymal")
  ) |>
  left_join(sample_dict, by = "brain")

thios_1 <- thios_all |> filter(slide == 1)
thios_2 <- thios_all |> filter(slide == 2)
```

```{r}
#| label: load-seurat
# Load pre-saved microglia Seurat objects (one per slide)
path <- "/Users/arbones/Dropbox/SyncBriefcase/LAB/UK/Xenium/ARIA_Akhil/aria_analysis/seurat_objects/"

slide1_mg <- qs2::qs_read(paste0(path, "20260225-mg_kai_slide1.qs2"))
slide2_mg <- qs2::qs_read(paste0(path, "20260225-mg_kai_slide2.qs2"))
```

## Compute distances

```{r}
#| label: compute-distances
# Convert HALO pixel coords to um centroids (factor = 0.2125)
make_centroids <- function(thios) {
  thios |>
    mutate(
      x_plaque = ((XMin + XMax) / 2) * 0.2125,
      y_plaque = ((YMin + YMax) / 2) * 0.2125,
      plaque_id = index
    ) |>
    select(plaque_id, brain, sample, treatment, plaque_type, x_plaque, y_plaque)
}

plaque_centroids_1 <- make_centroids(thios_1)
plaque_centroids_2 <- make_centroids(thios_2)

# Split centroids by plaque type
caa_coords_1   <- plaque_centroids_1 |> filter(plaque_type == "CAA")
caa_coords_2   <- plaque_centroids_2 |> filter(plaque_type == "CAA")
paren_coords_1 <- plaque_centroids_1 |> filter(plaque_type == "Parenchymal")
paren_coords_2 <- plaque_centroids_2 |> filter(plaque_type == "Parenchymal")

# k-d tree nearest-neighbour distance from each cell to nearest CAA and parenchymal plaque
compute_cell_distances <- function(seurat_obj, caa_coords, paren_coords) {
  coords_df <- GetTissueCoordinates(seurat_obj) |>
    as.data.frame() |>
    dplyr::select(cell, x, y)

  cell_mat  <- coords_df |> dplyr::select(x, y) |> as.matrix()
  caa_mat   <- caa_coords   |> dplyr::select(x_plaque, y_plaque) |> as.matrix()
  paren_mat <- paren_coords |> dplyr::select(x_plaque, y_plaque) |> as.matrix()

  caa_nn   <- nn2(data = caa_mat,   query = cell_mat, k = 1)
  paren_nn <- nn2(data = paren_mat, query = cell_mat, k = 1)

  coords_df |>
    mutate(
      dist_caa         = caa_nn$nn.dist[, 1],
      nearest_caa_id   = caa_coords$plaque_id[caa_nn$nn.idx[, 1]],
      dist_paren       = paren_nn$nn.dist[, 1],
      nearest_paren_id = paren_coords$plaque_id[paren_nn$nn.idx[, 1]]
    )
}

cell_distances_1 <- compute_cell_distances(slide1_mg, caa_coords_1, paren_coords_1)
cell_distances_2 <- compute_cell_distances(slide2_mg, caa_coords_2, paren_coords_2)
```

## Classify microglia as proximal or distal

 Microglia were classified as plaque-proximal (< 50 µm) or distal (≥ 50 µm) relative to CAA and     parenchymal plaques separately.

```{r}
#| label: add-proximity-metadata
# Add prox_caa, prox_paren, prox_any columns to Seurat metadata
add_proximity <- function(seurat_obj, cell_distances) {
  prox <- cell_distances |>
    mutate(
      prox_caa   = if_else(dist_caa < 50,   "proximal", "distal"),
      prox_paren = if_else(dist_paren < 50,  "proximal", "distal"),
      prox_any   = if_else(pmin(dist_caa, dist_paren) < 50, "proximal", "distal")
    ) |>
    column_to_rownames("cell") |>
    select(dist_caa, dist_paren, prox_caa, prox_paren, prox_any)

  AddMetaData(seurat_obj, prox)
}

slide1_mg <- add_proximity(slide1_mg, cell_distances_1)
slide2_mg <- add_proximity(slide2_mg, cell_distances_2)
```

## Cell counts per group

```{r}
#| label: cell-counts
# Proximal/distal counts per sample
bind_rows(slide1_mg@meta.data, slide2_mg@meta.data) |>
  group_by(sample_ID, prox_any) |>
  summarise(n = n(), .groups = "drop") |>
  pivot_wider(names_from = prox_any, values_from = n, values_fill = 0) |>
  mutate(total = proximal + distal) |>
  kbl(caption = "Microglia per sample — proximity to any plaque (50 um threshold)") |>
  kable_styling("striped", full_width = FALSE)
```

```{r}
#| label: pool-distances
# Pool per-slide distances into a single table with treatment annotation
cell_distances_all <- bind_rows(
  cell_distances_1 |> mutate(slide = 1L),
  cell_distances_2 |> mutate(slide = 2L)
) |>
  mutate(
    dist_any   = pmin(dist_caa, dist_paren),
    prox_caa   = if_else(dist_caa < 50,   "proximal", "distal"),
    prox_paren = if_else(dist_paren < 50,  "proximal", "distal"),
    prox_any   = if_else(dist_any < 50,    "proximal", "distal")
  )

# Join treatment info from Seurat metadata
treatment_lookup <- bind_rows(
  slide1_mg@meta.data |> as_tibble(rownames = "cell") |> select(cell, sample_ID, Treatment.Group),
  slide2_mg@meta.data |> as_tibble(rownames = "cell") |> select(cell, sample_ID, Treatment.Group)
)

cell_distances_all <- cell_distances_all |>
  left_join(treatment_lookup, by = "cell")
```

```{r}
#| label: ridge-distances-treatment
#| fig-width: 12
#| fig-height: 5
#| fig-cap: "Distribution of microglia-to-plaque distances by treatment group. Distances are capped at 500 um for visualization."

# Reshape to long format for faceted ridge plot
ridge_df <- cell_distances_all |>
  select(cell, Treatment.Group, dist_caa, dist_paren) |>
  pivot_longer(cols = c(dist_caa, dist_paren),
               names_to = "plaque_type", values_to = "distance") |>
  mutate(
    plaque_type  = recode(plaque_type, dist_caa = "CAA", dist_paren = "Parenchymal"),
    distance_cap = pmin(distance, 500)
  )

ggplot(ridge_df, aes(x = distance_cap, y = Treatment.Group, fill = Treatment.Group)) +
  geom_density_ridges(alpha = 0.6, scale = 1.2, rel_min_height = 0.01) +
  geom_vline(xintercept = 50, linetype = "dashed", colour = "grey30") +
  facet_wrap(~ plaque_type) +
  scale_fill_manual(values = c("IgG" = "#4DAF4A", "Adu" = "#E41A1C")) +
  labs(
    x = "Distance to nearest plaque (um)",
    y = NULL,
    fill = "Treatment"
  ) +
  theme_minimal(base_size = 15) +
  theme(legend.position = "bottom",
    strip.text = element_text(size = 18))
```

Ridge density plots of microglia-to-plaque distances revealed a treatment-dependent shift for CAA but not parenchymal plaques: in aducanumab-treated brains, microglia were distributed further from CAA plaques compared to IgG controls. In contrast, microglia-to-parenchymal plaque distance distributions were nearly identical across treatment groups, with a sharp peak within 50 µm reflecting dense plaque-associated microglia clustering. 

# Pseudobulk DE: Adu vs IgG in plaque-proximal microglia

Within microglia that sit < 50 um from a plaque, we test which genes differ
between aducanumab-treated (Adu, n = 3) and IgG-treated (n = 3) brains.
Pseudobulk counts are aggregated per brain, then DESeq2 tests Adu vs IgG.
This is run separately for CAA-proximal and parenchymal-proximal microglia.

```{r}
#| label: define-pseudobulk-builder
# Aggregate single-cell counts into pseudobulk profiles via sparse indicator matrix
build_pseudobulk <- function(slide_list, group_col) {
  # Extract raw count matrices from each slide
  all_counts <- map(slide_list, \(obj) {
    LayerData(obj, assay = DefaultAssay(obj), layer = "counts")
  })

  # Build metadata with grouping column
  all_meta <- map(slide_list, \(obj) {
    obj@meta.data |>
      rownames_to_column("cell") |>
      mutate(pb_group = .data[[group_col]])
  })

  # Keep only genes shared across slides, then column-bind
  shared_genes <- Reduce(intersect, map(all_counts, rownames))
  counts <- do.call(cbind, map(all_counts, \(m) m[shared_genes, ]))
  meta   <- bind_rows(all_meta)

  # Sparse indicator matrix: cells x groups, multiplied to sum counts per group
  groups <- unique(meta$pb_group)
  indicator <- Matrix::sparseMatrix(
    i = match(meta$cell, colnames(counts)),
    j = match(meta$pb_group, groups),
    x = 1,
    dims = c(ncol(counts), length(groups)),
    dimnames = list(colnames(counts), groups)
  )

  list(
    pb_counts = as.matrix(counts %*% indicator),
    meta      = meta,
    groups    = groups
  )
}
```


```{r}
#| label: define-de-function
# DESeq2 runner: fits model, extracts results, builds volcano + sig table
run_deseq2 <- function(
    pb_counts,
    col_data,
    design_formula,
    contrast,
    plot_title    = "DE results",
    plot_subtitle = "",
    pval_threshold = 0.01
) {
  # Build DESeq2 dataset and filter lowly-expressed genes (< 10 total counts)
  dds <- DESeqDataSetFromMatrix(
    countData = pb_counts,
    colData   = col_data,
    design    = design_formula
  )
  keep <- rowSums(counts(dds)) >= 10
  dds  <- dds[keep, ]

  # Fit DESeq2 model and extract results for the specified contrast
  dds <- DESeq(dds)
  res <- results(dds, contrast = contrast)

  # Tidy results table
  res_df <- as.data.frame(res) |>
    rownames_to_column("gene") |>
    as_tibble() |>
    dplyr::rename(
      p_val      = pvalue,
      p_val_adj  = padj,
      avg_log2FC = log2FoldChange
    ) |>
    arrange(p_val_adj)

  # Significant genes based on raw p-value threshold
  sig_df    <- res_df |> filter(p_val < pval_threshold)
  ev_df     <- res_df |> column_to_rownames("gene")
  sig_genes <- sig_df$gene

  # Colour significant genes red, rest grey
  keyvals <- rep("grey60", nrow(ev_df))
  names(keyvals) <- rownames(ev_df)
  keyvals[sig_genes] <- "red2"

  fc_range  <- range(ev_df$avg_log2FC, na.rm = TRUE)
  xlim_vals <- c(floor(fc_range[1]), ceiling(fc_range[2]))

  # Build volcano plot (EnhancedVolcano built-in cutoffs disabled)
  volcano <- EnhancedVolcano(
    ev_df,
    lab            = rownames(ev_df),
    selectLab      = sig_genes,
    x              = "avg_log2FC",
    y              = "p_val",
    ylim           = c(0, 3),
    title          = plot_title,
    subtitle       = plot_subtitle,
    labFace        = "italic",
    labSize        = 6,
    caption        = "",
    pCutoff        = 1,
    FCcutoff       = 0.001,
    colCustom      = keyvals,
    legendPosition = "none",
    drawConnectors = TRUE,
    directionConnectors = "both",
    max.overlaps   = Inf,
    xlim           = xlim_vals
  )

  # Add dashed line at the raw p-value of the least-significant DEG
  if (length(sig_genes) > 0) {
    raw_boundary <- max(ev_df[sig_genes, "p_val"], na.rm = TRUE)
    volcano <- volcano +
      geom_hline(yintercept = -log10(raw_boundary),
                 linetype = "dashed", colour = "black")
  }

  list(
    all_results         = ev_df,
    significant_results = sig_df |> column_to_rownames("gene"),
    volcano_plot        = volcano,
    sig_genes           = sig_genes,
    summary = list(
      total_genes       = nrow(ev_df),
      significant_genes = nrow(sig_df)
    )
  )
}
```

```{r}
#| label: define-treatment-de
# Subset to proximal (or distal) cells, pseudobulk by sample, test Adu vs IgG
run_treatment_de <- function(
    slide_list,
    proximity_col,
    proximity_value,
    plot_title
) {
  # Keep only cells matching the requested proximity group
  slide_list_sub <- map(slide_list, \(obj) {
    subset(obj, cells = colnames(obj)[obj[[proximity_col, drop = TRUE]] == proximity_value])
  })

  # Pseudobulk aggregation: one profile per sample (brain)
  pb <- build_pseudobulk(slide_list_sub, "sample_ID")

  # Map sample_ID to treatment group
  treatment_map <- bind_rows(
    map(slide_list, \(obj) {
      obj@meta.data |> distinct(sample_ID, Treatment.Group)
    })
  ) |> distinct()

  col_data <- data.frame(
    treatment = factor(
      treatment_map$Treatment.Group[match(pb$groups, treatment_map$sample_ID)],
      levels = c("IgG", "Adu")
    ),
    row.names = pb$groups
  )

  run_deseq2(pb$pb_counts, col_data,
             design_formula = ~ treatment,
             contrast       = c("treatment", "Adu", "IgG"),
             plot_title     = plot_title,
             plot_subtitle  = "Adu vs IgG")
}
```


The Xenium panel is small (480 genes) and not all genes are expressed in
microglia. After subsetting to proximal cells and pseudobulk aggregation across
only 6 samples (3 Adu + 3 IgG), many genes have near-zero total counts and are
dropped by the `rowSums >= 10` filter. The table below traces the number of
genes at each step.


Given the limited biological replication (n = 3 per group) and the restricted gene panel, no genes reached significance after BH correction. **We report genes with nominal p < 0.01** as exploratory candidates requiring validation.
```{r}
#| label: gene-funnel
# Step 1: genes in the Xenium panel (shared across slides)
panel_genes <- intersect(
  rownames(LayerData(slide1_mg, layer = "counts")),
  rownames(LayerData(slide2_mg, layer = "counts"))
)

# Step 2: genes detected in at least 1 microglia cell (any proximity)
all_counts <- cbind(
  LayerData(slide1_mg, layer = "counts")[panel_genes, ],
  LayerData(slide2_mg, layer = "counts")[panel_genes, ]
)
detected_in_mg <- sum(Matrix::rowSums(all_counts > 0) > 0)

# Step 3–4: genes surviving pseudobulk count filter per proximity group
count_after_pb <- function(slide_list, proximity_col, proximity_value) {
  sub <- map(slide_list, \(obj) {
    subset(obj, cells = colnames(obj)[obj[[proximity_col, drop = TRUE]] == proximity_value])
  })
  pb <- build_pseudobulk(sub, "sample_ID")
  n_cells <- sum(map_int(sub, ncol))
  n_genes_before <- nrow(pb$pb_counts)
  n_genes_after  <- sum(rowSums(pb$pb_counts) >= 10)
  tibble(n_cells = n_cells, genes_in_pb = n_genes_before, genes_after_filter = n_genes_after)
}

caa_stats   <- count_after_pb(list(slide1_mg, slide2_mg), "prox_caa",   "proximal")
paren_stats <- count_after_pb(list(slide1_mg, slide2_mg), "prox_paren", "proximal")

tibble(
  step = c(
    "Xenium panel (shared across slides)",
    "Detected in >=1 microglia cell",
    "CAA-proximal: cells in pseudobulk",
    "CAA-proximal: genes in pseudobulk",
    "CAA-proximal: genes after >=10 counts filter",
    "Paren-proximal: cells in pseudobulk",
    "Paren-proximal: genes in pseudobulk",
    "Paren-proximal: genes after >=10 counts filter"
  ),
  count = c(
    length(panel_genes),
    detected_in_mg,
    caa_stats$n_cells,
    caa_stats$genes_in_pb,
    caa_stats$genes_after_filter,
    paren_stats$n_cells,
    paren_stats$genes_in_pb,
    paren_stats$genes_after_filter
  )
) |>
  kbl(caption = "Gene filtering funnel — from Xenium panel to testable genes in DESeq2") |>
  kable_styling("striped", full_width = FALSE)
```

## CAA — treatment DE

```{r}
#| label: trt-caa-proximal
#| fig-width: 14
#| fig-height: 6
trt_caa_prox <- run_treatment_de(
  list(slide1_mg, slide2_mg), "prox_caa", "proximal",
  "Microglia proximal to CAA — Adu vs IgG"
)

# Significant genes table (or placeholder if none)
sig_tbl <- if (nrow(trt_caa_prox$significant_results) > 0) {
  trt_caa_prox$significant_results |>
    dplyr::select(p_val, avg_log2FC, p_val_adj) |>
    mutate(across(where(is.numeric), \(x) signif(x, 3))) |>
    arrange(p_val_adj) |>
    tableGrob(theme = ttheme_minimal(base_size = 14))
} else {
  tableGrob(data.frame(Result = "No significant DEGs"), theme = ttheme_minimal(base_size = 16))
}

trt_caa_prox$volcano_plot + wrap_elements(sig_tbl) + plot_layout(widths = c(2, 1))
```

## Parenchymal — treatment DE

```{r}
#| label: trt-paren-proximal
#| fig-width: 14
#| fig-height: 6
trt_paren_prox <- run_treatment_de(
  list(slide1_mg, slide2_mg), "prox_paren", "proximal",
  "Microglia proximal to parenchymal — Adu vs IgG"
)

sig_tbl <- if (nrow(trt_paren_prox$significant_results) > 0) {
  trt_paren_prox$significant_results |>
    dplyr::select(p_val, avg_log2FC, p_val_adj) |>
    mutate(across(where(is.numeric), \(x) signif(x, 3))) |>
    arrange(p_val_adj) |>
    tableGrob(theme = ttheme_minimal(base_size = 14))
} else {
  tableGrob(data.frame(Result = "No significant DEGs"), theme = ttheme_minimal(base_size = 16))
}

trt_paren_prox$volcano_plot + wrap_elements(sig_tbl) + plot_layout(widths = c(2, 1))
```

Aducanumab promotes a plaque-proximal microglial state marked by chemokine and inflammasome priming (_Ccl12, Ifitm3, Nlrp3_) and selective metabolic remodeling (Hk3↑, Pfkm↓, Ldhb↓), without engagement of the canonical _Itgax_⁺ DAM transcriptional program