---
title: "Multi-plaque neighbourhood effects"
subtitle: "Disentangling plaque proximity from local plaque density"
date: last-modified
execute:
  echo: true
  warning: false
embed-resources: true
format: html
code-fold: true
tags: [xenium, coregistration, halo, density, neighbourhood]
---

# Setup and data loading

```{r}
#| label: setup
source("R/setup.R")
library(ggrepel)
library(patchwork)
library(broom)
library(ggridges)

data <- load_slides_and_distances("full")
slide1 <- data$slide1; slide2 <- data$slide2
cell_distances_1 <- data$distances_1; cell_distances_2 <- data$distances_2
cell_distances_all <- data$distances_all
```

# Plaque density computation

The existing pipeline assigns each cell to its single nearest plaque. But do cells surrounded by **many** plaques behave differently than cells near an isolated plaque? We compute, for each cell, the number of plaques of each type within concentric radii (50, 100, 150 µm).

```{r}
#| label: compute-plaque-density
compute_plaque_density <- function(cell_coords, plaque_coords, radii = c(50, 100, 150)) {
  cell_mat   <- cell_coords |> dplyr::select(x, y) |> as.matrix()
  plaque_mat <- plaque_coords |> dplyr::select(x_plaque, y_plaque) |> as.matrix()
  k_use      <- min(nrow(plaque_mat), 50)
  nn_res     <- nn2(data = plaque_mat, query = cell_mat, k = k_use)

  density_cols <- map(radii, \(r) rowSums(nn_res$nn.dists <= r)) |>
    set_names(paste0("n_plaques_", radii, "um"))

  bind_cols(cell_coords |> dplyr::select(cell), as_tibble(density_cols))
}

# Compute density for each plaque type and all plaques, per slide
compute_all_densities <- function(cell_dist, caa_coords, paren_coords, all_coords) {
  cell_xy <- cell_dist |> dplyr::select(cell, x, y)

  dens_caa   <- compute_plaque_density(cell_xy, caa_coords) |>
    dplyr::rename(n_caa_50um = n_plaques_50um, n_caa_100um = n_plaques_100um, n_caa_150um = n_plaques_150um)

  dens_paren <- compute_plaque_density(cell_xy, paren_coords) |>
    dplyr::rename(n_paren_50um = n_plaques_50um, n_paren_100um = n_plaques_100um, n_paren_150um = n_plaques_150um)

  dens_any   <- compute_plaque_density(cell_xy, all_coords) |>
    dplyr::rename(n_any_50um = n_plaques_50um, n_any_100um = n_plaques_100um, n_any_150um = n_plaques_150um)

  cell_dist |>
    left_join(dens_caa,   by = "cell") |>
    left_join(dens_paren, by = "cell") |>
    left_join(dens_any,   by = "cell")
}

cell_distances_1 <- compute_all_densities(
  cell_distances_1, caa_coords_1, paren_coords_1, plaque_centroids_1
)
cell_distances_2 <- compute_all_densities(
  cell_distances_2, caa_coords_2, paren_coords_2, plaque_centroids_2
)

# Re-pool with density columns
cell_distances_all <- bind_rows(
  cell_distances_1 |> mutate(slide = 1L),
  cell_distances_2 |> mutate(slide = 2L)
) |>
  mutate(dist_any = pmin(dist_caa, dist_paren)) |>
  left_join(treatment_lookup, by = "cell")
```

## Density summary

```{r}
#| label: density-summary-table
cell_distances_all |>
  group_by(sample_ID, Treatment.Group) |>
  summarise(
    n_cells           = n(),
    mean_n_any_100um  = mean(n_any_100um),
    median_n_any_100um = median(n_any_100um),
    sd_n_any_100um    = sd(n_any_100um),
    mean_n_caa_100um  = mean(n_caa_100um),
    mean_n_paren_100um = mean(n_paren_100um),
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric) & !c(n_cells), \(x) round(x, 2))) |>
  kbl(caption = "Plaque density per sample (100 µm radius)") |>
  kable_styling("striped", full_width = FALSE)
```

**Results:** The density summary table reveals substantial variation in local plaque burden across samples. IgG-treated brain KK4_465 has the highest mean plaque density (7.04 plaques within 100 µm per cell, median = 5), while Adu-treated KK4_492 has the lowest (mean = 3.68, median = 2). Notably, the two treatment groups do not show a clean separation: IgG samples tend toward higher density (means 4.86--7.04) compared to Adu samples (means 3.68--5.94), but inter-brain variability within each group is considerable. This sample-level heterogeneity motivates the per-brain consistency checks performed later. The standard deviations are large relative to the means (coefficients of variation ~0.9--1.1), confirming that plaque density is highly spatially heterogeneous within each brain section --- some tissue regions are plaque-dense while others are relatively spared.

## Density distributions

```{r}
#| label: density-distribution-plots
#| fig-height: 5
#| fig-width: 10
p_hist <- cell_distances_all |>
  ggplot(aes(x = n_any_100um, fill = Treatment.Group)) +
  geom_histogram(binwidth = 1, position = "dodge", alpha = 0.7) +
  scale_fill_manual(values = c(Adu = "#E07B39", IgG = "#7B68AE")) +
  labs(x = "Number of plaques within 100 µm", y = "Cell count", fill = "Treatment") +
  theme_minimal(base_size = 13)

p_hex <- cell_distances_all |>
  ggplot(aes(x = dist_any, y = n_any_100um)) +
  geom_hex(bins = 80) +
  scale_fill_viridis_c(trans = "log10") +
  labs(x = "Distance to nearest plaque (µm)", y = "Plaques within 100 µm") +
  theme_minimal(base_size = 13)

p_hist + p_hex
```

**Results:** The histogram (left panel) shows that both Adu and IgG cells follow a strongly right-skewed distribution: the majority of cells have 0--5 plaques within 100 µm, with a long tail extending past 40. IgG cells (purple) are slightly over-represented at higher density values, consistent with the higher mean densities seen in the summary table. The hexbin plot (right panel) reveals the expected inverse relationship between distance to nearest plaque and local plaque count: cells very close to a plaque (distance < 50 µm) tend to have more plaques in their neighbourhood, while distant cells (> 200 µm) almost universally have zero local plaques. However, the relationship is not strictly monotonic --- some cells at intermediate distances (50--100 µm) still sit in high-density regions, which is the key insight motivating this analysis: distance to the nearest plaque and the number of surrounding plaques are correlated but distinct features of the local microenvironment.

# Isolated vs clustered plaque environments

Cells are classified as **isolated** (0–1 plaques within 100 µm) or **clustered** (≥2 plaques within 100 µm). This binary split captures whether a cell sits in a sparse or dense plaque environment.

```{r}
#| label: stratify-isolated-vs-clustered
cell_distances_all <- cell_distances_all |>
  mutate(
    density_class = if_else(n_any_100um <= 1, "isolated", "clustered"),
    density_class = factor(density_class, levels = c("isolated", "clustered"))
  )

cell_distances_all |>
  count(Treatment.Group, density_class) |>
  group_by(Treatment.Group) |>
  mutate(pct = round(n / sum(n) * 100, 1)) |>
  ungroup() |>
  kbl(caption = "Cells by density class and treatment") |>
  kable_styling("striped", full_width = FALSE)
```

**Results:** Across both treatment groups, roughly two-thirds of cells reside in clustered environments (64.9% Adu, 70.3% IgG) and one-third in isolated environments. The slightly higher proportion of clustered cells in IgG animals is consistent with the higher mean plaque densities observed in IgG samples. Importantly, both classes contain tens of thousands of cells per treatment (>50k isolated, >115k clustered), ensuring adequate statistical power for downstream comparisons within each stratum.

## Spatial map of plaque density

**Concept and calculation.** The spatial map visualises every Xenium-profiled cell as a point in tissue space (x and y coordinates in micrometres), coloured by its **local plaque density** --- defined as the number of ThioS-positive plaques (of any type: CAA or parenchymal) whose centroid falls within a 100 µm radius of that cell. This metric was computed using a k-d tree nearest-neighbour search (`RANN::nn2` with k = 50), counting how many of the 50 nearest plaque centroids lie within the 100 µm threshold. The colour scale uses the "inferno" palette from viridis, where dark/black regions indicate cells with zero or very few nearby plaques and bright yellow/white regions indicate cells surrounded by many plaques (up to ~50). White dots overlaid on the map mark plaque centroid positions themselves. Each of the six panels corresponds to one sagittal brain section (sample_ID), arranged in a 2-row by 3-column grid (top row: KK4_464, KK4_465, KK4_492; bottom row: KK4_496, KK4_502, KK4_504). Within each panel, the five spatially disconnected tissue islands correspond to the five anatomical regions annotated in the Seurat `Region` metadata variable: **Cortex**, **Hippocampus**, **Hypothalamus**, **Thalamus**, and **WM** (white matter). These regions appear as separate clusters because they are physically discontinuous structures within a single sagittal section. Axis scales are free (`scales = "free"` in `facet_wrap`), meaning each panel's x and y ranges adjust independently to the spatial extent of that brain section. This is necessary because each tissue section occupies a different footprint in the Xenium field of view; a shared axis range would shrink smaller sections into unusable thumbnails.

**Results.** The spatial maps reveal that plaque density is not uniformly distributed across the tissue but instead forms distinct **hotspot clusters** --- localised regions where plaques aggregate and cells experience high amyloid burden. These hotspots appear as bright yellow patches, typically spanning several hundred micrometres, and they correspond to anatomical regions with heavy ThioS staining. Surrounding these hotspots are broad dark regions where cells are far from any plaque and the local density drops to zero.

Several patterns emerge across the six brains:

- **KK4_465 (IgG)** shows the most extensive high-density regions, consistent with it having the highest mean plaque density (7.04) in the summary table. The bright patches are widespread across the cortex and hippocampal regions.
- **KK4_492 (Adu)** has fewer and smaller hotspots, consistent with its low mean density (3.68). This sample shows the clearest spatial segregation between plaque-rich and plaque-poor zones.
- **KK4_502 (Adu)** and **KK4_496 (IgG)** show intermediate patterns with several discrete high-density foci.
- The white plaque centroid markers confirm that bright cell regions co-localise with actual plaque positions, validating the density metric.

The maps highlight a critical confound for gene-distance analyses: a cell 30 µm from a single isolated plaque experiences a very different amyloid microenvironment than a cell 30 µm from one plaque but also within 100 µm of five additional plaques. The former sits in an "isolated" niche; the latter is immersed in a confluent plaque region. This spatial heterogeneity is exactly what the density stratification in subsequent sections aims to disentangle.

```{r}
#| label: density-spatial-map
#| fig-height: 10
#| fig-width: 12
cell_distances_all |>
  ggplot(aes(x = x, y = y, colour = n_any_100um)) +
  geom_point(size = 0.1, alpha = 0.3) +
  scale_colour_viridis_c(option = "inferno", name = "Plaques\n(100 µm)") +
  geom_point(
    data = bind_rows(plaque_centroids_1, plaque_centroids_2),
    aes(x = x_plaque, y = y_plaque),
    colour = "white", size = 0.3, alpha = 0.5, inherit.aes = FALSE
  ) +
  facet_wrap(~sample_ID, scales = "free", ncol = 3) +
  labs(title = "Local plaque density per cell", x = "x (µm)", y = "y (µm)") +
  theme_dark(base_size = 12) +
  theme(legend.position = "right")
```

# Gene expression by density class

**Concept and calculation.** Disease-associated microglia (DAM) represent an activated microglial state first described in amyloid-bearing mouse models, characterised by upregulation of a specific gene programme including *Cst7*, *Trem2*, *Cd68*, *Lyz2*, *Bcl2a1b*, and *Il1b*. These genes are known markers of the microglial response to amyloid-beta plaques. Here we ask: do cells in **clustered** plaque environments (surrounded by 2 or more plaques within 100 µm) show higher DAM gene expression than cells in **isolated** environments (0--1 plaque within 100 µm)?

Raw transcript counts for each DAM gene were extracted from the Xenium count matrices across both slides, then joined to the density classification computed above. Violin plots display the full distribution of raw counts for each gene, split by density class, with the median marked by a horizontal line within each violin. This is a simple marginal comparison --- it does not control for distance to the nearest plaque, which the interaction models below address.

Do DAM-associated genes show higher expression in clustered plaque environments?

```{r}
#| label: gene-expression-by-density
#| fig-height: 8
#| fig-width: 10
dam_genes <- c("Cst7", "Trem2", "Cd68", "Il1b", "Lyz2", "Bcl2a1b")

# Extract counts for DAM genes across both slides
extract_gene_counts <- function(seurat_list, genes) {
  map(seurat_list, \(obj) {
    available <- intersect(genes, rownames(obj))
    counts <- LayerData(obj, assay = DefaultAssay(obj), layer = "counts")[available, ]
    as.matrix(t(counts)) |>
      as.data.frame() |>
      rownames_to_column("cell")
  }) |>
    bind_rows()
}

gene_counts <- extract_gene_counts(list(slide1, slide2), dam_genes)

expr_density <- cell_distances_all |>
  inner_join(gene_counts, by = "cell") |>
  pivot_longer(
    cols = any_of(dam_genes),
    names_to = "gene",
    values_to = "count"
  )

expr_density |>
  ggplot(aes(x = density_class, y = count, fill = density_class)) +
  geom_violin(scale = "width", alpha = 0.7, draw_quantiles = 0.5) +
  scale_fill_manual(values = c(isolated = "grey70", clustered = "#E07B39")) +
  facet_wrap(~gene, scales = "free_y", ncol = 3) +
  labs(
    title = "DAM gene expression: isolated vs clustered plaque environments",
    x = NULL, y = "Raw count", fill = "Density class"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

**Results.** The violin plots reveal a consistent pattern: all six DAM-associated genes show visibly higher expression in the **clustered** class (orange) compared to the **isolated** class (grey). The effect is most pronounced for *Cst7* (maximum raw counts reaching ~90 in clustered vs ~60 in isolated), *Lyz2* (counts exceeding 100 in clustered), and *Trem2* (counts up to ~40 in clustered). *Cd68*, *Bcl2a1b*, and *Il1b* show similar trends but with lower absolute counts, consistent with their generally sparser expression in the Xenium panel.

Importantly, both density classes show the characteristic zero-inflated distribution typical of spatial transcriptomics data: the vast majority of cells express zero copies of any given DAM gene, with a small proportion showing non-zero expression. The violin shapes therefore appear as thin lines with bulges at specific count values. The key observation is that the bulges at higher count values are larger and extend further in the clustered class, indicating that cells surrounded by multiple plaques are not only more likely to express DAM genes but also express them at higher levels when they do.

This marginal comparison, however, conflates two effects: (1) cells in clustered environments tend to be closer to their nearest plaque (as shown in the hexbin plot), and (2) cells may respond more strongly to plaque proximity when surrounded by additional plaques. The interaction models below formally disentangle these two contributions.

# Distance × density interaction models

Does the gene-distance relationship depend on local plaque density? We fit Poisson GLMs of the form `count ~ dist_any * n_any_100um` and test whether the interaction term is significant via a likelihood-ratio test.

**Caveats:**

- Poisson GLM may underestimate standard errors under overdispersion; negative binomial (`MASS::glm.nb()`) is used as robustness check below.
- Cell-level analysis introduces pseudoreplication (n = 3 vs 3 brains). Per-brain consistency (next section) partially addresses this.
- With ~360k cells, even tiny effects reach significance — we emphasise effect sizes over p-values.

```{r}
#| label: model-distance-density-interaction
# Get all expressed genes
count_list <- map(list(slide1, slide2), \(obj) {
  LayerData(obj, assay = DefaultAssay(obj), layer = "counts")
})
shared_genes <- Reduce(intersect, map(count_list, rownames))
counts_all <- do.call(cbind, map(count_list, \(m) m[shared_genes, ]))

shared_cells <- intersect(colnames(counts_all), cell_distances_all$cell)
counts_all <- counts_all[, shared_cells]
n_expressing <- Matrix::rowSums(counts_all > 0)
genes_to_test <- names(n_expressing[n_expressing >= 100])

model_df <- cell_distances_all |>
  filter(cell %in% shared_cells) |>
  arrange(match(cell, shared_cells))

fit_density_model <- function(gene_name) {
  expr_vec <- as.numeric(counts_all[gene_name, ])
  df <- tibble(
    count       = expr_vec,
    dist_any    = model_df$dist_any,
    n_any_100um = model_df$n_any_100um
  )

  m_add <- glm(count ~ dist_any + n_any_100um, family = poisson(link = "log"), data = df)
  m_int <- glm(count ~ dist_any * n_any_100um, family = poisson(link = "log"), data = df)
  lrt   <- anova(m_add, m_int, test = "Chisq")

  int_coef <- tidy(m_int) |> filter(term == "dist_any:n_any_100um")

  tibble(
    gene          = gene_name,
    int_estimate  = int_coef$estimate,
    int_se        = int_coef$std.error,
    int_z         = int_coef$statistic,
    int_pval      = int_coef$p.value,
    lrt_pval      = lrt$`Pr(>Chi)`[2],
    aic_additive  = AIC(m_add),
    aic_interact  = AIC(m_int),
    delta_aic     = AIC(m_add) - AIC(m_int)
  )
}

interaction_results <- map(genes_to_test, fit_density_model, .progress = "GLM interactions") |>
  bind_rows() |>
  mutate(
    padj = p.adjust(lrt_pval, method = "BH"),
    sig  = padj < 0.05
  ) |>
  arrange(lrt_pval)

interaction_results |>
  slice_head(n = 20) |>
  mutate(across(where(is.numeric), \(x) signif(x, 3))) |>
  kbl(caption = "Top 20 genes with distance × density interaction (Poisson GLM)") |>
  kable_styling("striped", full_width = FALSE) |>
  scroll_box(width = "100%")
```

**Results:** Of the 459 genes tested, a striking 437 (95%) show a significant distance x density interaction after BH correction (padj < 0.05). This near-universal significance reflects the enormous sample size (~358k cells) rather than necessarily large biological effects --- as noted in the caveats, even tiny coefficients become significant with hundreds of thousands of observations. The median interaction coefficient among significant genes is 0.000129, which is small in absolute terms but interpretable on the log-link scale of the Poisson GLM.

Among the top 20 genes, two classes of interaction emerge:

- **Positive interaction coefficients** (e.g., *App* = 0.000946, *Acsl4* = 0.000688, *Bace1* = 0.000734): These genes show a *steeper* increase in expression near plaques when local plaque density is high. In other words, the proximity effect is amplified by surrounding plaques. *App* (amyloid precursor protein) has the largest delta AIC (44,100), indicating the interaction model provides a dramatically better fit. *Bace1* (beta-secretase 1) is particularly notable as a key enzyme in amyloid-beta production.
- **Negative interaction coefficients** (e.g., *C1qa* = -0.00212, *Bcl2a1b* = -0.00246, *Bgn* = -0.00313, *Aif1* = -0.00172, *Axl* = -0.00168): These genes show a *weaker* proximity effect in high-density regions. This may seem counterintuitive, but can reflect a saturation or ceiling effect: in dense plaque environments, these genes are already maximally upregulated regardless of the distance to the single nearest plaque. *C1qa* (complement component) and *Aif1* (Iba1, a pan-microglial marker) are classical neuroinflammatory genes whose expression may plateau in heavily burdened regions.

## Radius sensitivity

Do the interaction effects depend on the radius used to define plaque density? We repeat the analysis at 50, 100, and 150 µm.

```{r}
#| label: radius-sensitivity
#| fig-height: 6
#| fig-width: 8
top_interact_genes <- interaction_results |>
  filter(sig) |>
  slice_head(n = 20) |>
  pull(gene)

fit_at_radius <- function(gene_name, radius_col) {
  expr_vec <- as.numeric(counts_all[gene_name, ])
  df <- tibble(
    count   = expr_vec,
    dist    = model_df$dist_any,
    density = model_df[[radius_col]]
  )
  m_int <- glm(count ~ dist * density, family = poisson(link = "log"), data = df)
  int_coef <- tidy(m_int) |> filter(term == "dist:density")
  tibble(gene = gene_name, radius = radius_col, estimate = int_coef$estimate, pval = int_coef$p.value)
}

radius_cols <- c("n_any_50um", "n_any_100um", "n_any_150um")

sensitivity_results <- expand_grid(gene = top_interact_genes, radius = radius_cols) |>
  pmap(\(gene, radius) fit_at_radius(gene, radius), .progress = "Radius sensitivity") |>
  bind_rows()

sensitivity_results |>
  mutate(
    radius_label = str_extract(radius, "\\d+") |> paste0(" µm"),
    sig_star = case_when(pval < 0.001 ~ "***", pval < 0.01 ~ "**", pval < 0.05 ~ "*", TRUE ~ "")
  ) |>
  ggplot(aes(x = radius_label, y = gene, fill = estimate)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sig_star), size = 4) +
  scale_fill_gradient2(low = "#7B68AE", mid = "white", high = "#E07B39", name = "Interaction\ncoefficient") +
  labs(
    title = "Distance × density interaction across radii",
    subtitle = "Stars: * p < 0.05, ** p < 0.01, *** p < 0.001",
    x = "Density radius", y = NULL
  ) +
  theme_minimal(base_size = 13)
```

**Results:** The heatmap confirms that all 20 top interaction genes are highly significant (*** at p < 0.001) across all three radii (50, 100, and 150 µm). The colour gradient shows that the sign and relative magnitude of the interaction coefficient are consistent across radii: genes with positive interactions at 100 µm (e.g., *App*, *Bin1*, *Bhlhe40*, warm colours) remain positive at 50 and 150 µm, and likewise for negative interactions (e.g., *C1qa*, *Bgn*, *Bcl2a1b*, cool colours). This radius-invariance suggests that the density effect is not an artifact of the arbitrary 100 µm threshold but reflects a genuine spatial property of the plaque microenvironment. The interaction coefficients tend to be slightly larger in absolute value at smaller radii (50 µm), which is expected: a tighter radius more precisely captures the immediate plaque neighbourhood and reduces dilution from distant, irrelevant plaques.

# Per-brain consistency

Cell-level GLMs treat each cell as independent, inflating significance. To check whether the interaction effect is consistent across biological replicates, we refit the model per brain and examine sign consistency.

```{r}
#| label: per-brain-consistency
#| fig-height: 6
#| fig-width: 10
brain_lookup <- treatment_lookup |>
  mutate(brain = str_extract(sample_ID, "KK4_\\d+"))

model_df_brain <- model_df |>
  left_join(
    brain_lookup |> dplyr::select(cell, brain_id = sample_ID),
    by = "cell"
  )

fit_per_brain <- function(gene_name, brain_id_val) {
  idx <- which(model_df_brain$brain_id == brain_id_val)
  if (length(idx) < 50) return(NULL)

  expr_vec <- as.numeric(counts_all[gene_name, idx])
  df <- tibble(
    count       = expr_vec,
    dist_any    = model_df_brain$dist_any[idx],
    n_any_100um = model_df_brain$n_any_100um[idx]
  )

  tryCatch({
    m <- glm(count ~ dist_any * n_any_100um, family = poisson(link = "log"), data = df)
    int_coef <- tidy(m) |> filter(term == "dist_any:n_any_100um")
    tibble(gene = gene_name, brain = brain_id_val, estimate = int_coef$estimate, pval = int_coef$p.value)
  }, error = \(e) NULL)
}

brains <- unique(model_df_brain$brain_id)
brains <- brains[!is.na(brains)]

per_brain_results <- expand_grid(gene = top_interact_genes, brain = brains) |>
  pmap(\(gene, brain) fit_per_brain(gene, brain), .progress = "Per-brain models") |>
  compact() |>
  bind_rows()

per_brain_results |>
  mutate(sign = if_else(estimate > 0, "+", "−")) |>
  ggplot(aes(x = brain, y = gene, fill = estimate)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sign), size = 5) +
  scale_fill_gradient2(low = "#7B68AE", mid = "white", high = "#E07B39", name = "Interaction\ncoefficient") +
  labs(
    title = "Per-brain interaction coefficients",
    subtitle = "Consistent sign across brains supports biological (not pseudoreplication) signal",
    x = "Brain (sample_ID)", y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Results:** The per-brain heatmap provides crucial biological validation. Of the 20 top interaction genes, the sign of the interaction coefficient is perfectly consistent across all 6 brains for the majority of genes. Genes like *App*, *Bin1*, *Bhlhe40*, *Bace1*, *Auh*, *Aspa*, *Acss2*, *Acsl6*, *Acsl4*, and *Acat1* show a positive (+) interaction in every single brain, regardless of treatment group. Conversely, *C1qa*, *Bgn*, *Bcl2a1b*, *B4galt1*, *Axl*, *Apod*, *Aldh2*, *Aif1*, and *Acta2* show a negative (-) interaction in all 6 brains. *B3gat3* also shows consistent positive sign across all brains.

This perfect sign consistency across 6 independent biological replicates (3 Adu + 3 IgG) strongly argues that the interaction effects are genuine biology, not pseudoreplication artifacts. If the cell-level significance were driven purely by the large sample size without an underlying biological signal, we would expect inconsistent signs across brains. The fact that 20 out of 20 tested genes show unanimous sign agreement is remarkably robust.

# Gene-distance correlation: isolated vs clustered

Re-run `compute_gene_distance_correlation()` separately for cells in isolated vs clustered plaque environments. If the same genes appear enriched near plaques in both strata, the distance effect is independent of density.

```{r}
#| label: density-class-gene-distance-correlation
compute_gene_distance_correlation <- function(
    seurat_list,
    cell_distances,
    distance_col = "dist_any",
    min_expressing_cells = 100
) {
  count_list <- map(seurat_list, \(obj) {
    LayerData(obj, assay = DefaultAssay(obj), layer = "counts")
  })

  shared_genes <- Reduce(intersect, map(count_list, rownames))
  counts <- do.call(cbind, map(count_list, \(m) m[shared_genes, ]))

  shared_cells <- intersect(colnames(counts), cell_distances$cell)
  counts <- counts[, shared_cells]
  dist_vec <- cell_distances |>
    filter(cell %in% shared_cells) |>
    arrange(match(cell, shared_cells)) |>
    pull(!!sym(distance_col))

  n_expressing <- Matrix::rowSums(counts > 0)
  keep <- n_expressing >= min_expressing_cells
  counts <- counts[keep, ]

  results <- map(seq_len(nrow(counts)), \(i) {
    gene_expr <- counts[i, ]
    ct <- cor.test(as.numeric(gene_expr), dist_vec, method = "spearman", exact = FALSE)
    tibble(gene = rownames(counts)[i], rho = ct$estimate, pval = ct$p.value)
  }, .progress = "Spearman correlations") |>
    bind_rows()

  results |>
    mutate(
      padj     = p.adjust(pval, method = "BH"),
      rho_flip = -rho,
      rank     = rank(-rho_flip, ties.method = "first"),
      sig      = case_when(
        padj < 0.05 & rho_flip > 0 ~ "enriched_near",
        padj < 0.05 & rho_flip < 0 ~ "enriched_far",
        TRUE                        ~ "ns"
      )
    ) |>
    arrange(rank)
}

seurat_list <- list(slide1, slide2)

# Split cells by density class
cells_isolated  <- cell_distances_all |> filter(density_class == "isolated")
cells_clustered <- cell_distances_all |> filter(density_class == "clustered")

cor_isolated <- compute_gene_distance_correlation(
  seurat_list, cells_isolated, distance_col = "dist_any"
)

cor_clustered <- compute_gene_distance_correlation(
  seurat_list, cells_clustered, distance_col = "dist_any"
)
```

```{r}
#| label: density-class-rank-plots
#| fig-height: 5
#| fig-width: 12
plot_gene_distance_correlation <- function(
    results,
    title          = "Gene-distance correlation",
    n_top          = 5,
    genes_to_label = NULL
) {
  top_pos <- results |> filter(sig == "enriched_near") |> slice_min(rank, n = n_top)
  top_neg <- results |> filter(sig == "enriched_far")  |> slice_max(rank, n = n_top)
  auto_label <- bind_rows(top_pos, top_neg)

  if (!is.null(genes_to_label)) {
    user_genes <- results |> filter(gene %in% genes_to_label)
    auto_label <- bind_rows(auto_label, user_genes) |> distinct(gene, .keep_all = TRUE)
  }

  colour_map <- c(
    "enriched_near" = "#E07B39",
    "enriched_far"  = "#7B68AE",
    "ns"            = "grey70"
  )

  ggplot(results, aes(x = rank, y = rho_flip)) +
    geom_point(aes(colour = sig), size = 2, alpha = 0.5) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
    geom_text_repel(
      data = auto_label, aes(label = gene, colour = sig),
      size = 4, fontface = "italic", box.padding = 0.4, max.overlaps = 20, show.legend = FALSE
    ) +
    scale_colour_manual(
      values = colour_map,
      labels = c(
        "enriched_near" = "Enriched near plaque",
        "enriched_far"  = "Enriched far from plaque",
        "ns"            = "Not significant"
      )
    ) +
    labs(title = title, x = "Gene rank", y = "Spearman rho (flipped)", colour = NULL) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank(), panel.grid.major.x = element_blank())
}

p_iso  <- plot_gene_distance_correlation(cor_isolated,  title = "Isolated (≤1 plaque within 100 µm)")
p_clust <- plot_gene_distance_correlation(cor_clustered, title = "Clustered (≥2 plaques within 100 µm)")

(p_iso | p_clust) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

```{r}
#| label: rank-comparison-scatter
#| fig-height: 6
#| fig-width: 6
rank_comparison <- cor_isolated |>
  dplyr::select(gene, rho_flip_isolated = rho_flip, sig_isolated = sig) |>
  inner_join(
    cor_clustered |> dplyr::select(gene, rho_flip_clustered = rho_flip, sig_clustered = sig),
    by = "gene"
  )

label_genes <- rank_comparison |>
  filter(gene %in% dam_genes | abs(rho_flip_isolated - rho_flip_clustered) > 0.05)

ggplot(rank_comparison, aes(x = rho_flip_isolated, y = rho_flip_clustered)) +
  geom_point(alpha = 0.4, colour = "grey50") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "red") +
  geom_text_repel(
    data = label_genes,
    aes(label = gene),
    size = 3.5, fontface = "italic", max.overlaps = 15
  ) +
  labs(
    title = "Gene-distance correlation: isolated vs clustered",
    subtitle = "Genes on diagonal have consistent effects regardless of density",
    x = "rho_flip (isolated)", y = "rho_flip (clustered)"
  ) +
  theme_minimal(base_size = 13) +
  coord_equal()
```

**Results:** The side-by-side rank plots reveal striking differences between the two microenvironments. In the **isolated** stratum (left panel), the gene-distance correlations are modest: the top enriched-near-plaque genes (*App*, *Snap25*, *Slc17a7*, *Tmsb4x*, *Grin2a*) have rho_flip values around 0.05--0.12, and the enriched-far genes (*Dnah11*, *Htr2c*, *Ttr*) reach only about -0.10 to -0.15. These are predominantly neuronal genes, not classical immune/DAM markers, suggesting that near an isolated plaque, the dominant spatial signal reflects neuronal tissue architecture rather than an immune response.

In the **clustered** stratum (right panel), the picture changes dramatically. The top enriched-near-plaque genes are now canonical DAM/immune markers: *Cst7* (rho_flip ~0.22), *Ctsd*, *Bcl2a1b*, *Lyz2*, *Trem2* --- all classic disease-associated microglia genes. The effect sizes are roughly double those in the isolated stratum (rho_flip up to 0.2 vs 0.1). Enriched-far genes include *Fdf1*, *Chst1*, *App* --- notably, *App* switches from being enriched near plaques in the isolated setting to being enriched far from plaques in the clustered setting.

The **scatter plot** comparing rho_flip values between strata quantifies this directly. Most genes fall near the diagonal (red dashed line), indicating that the majority of gene-distance relationships are consistent regardless of plaque density. However, notable outliers exist:

- Genes **above the diagonal** (e.g., *Cst7*, *Lyz2*, *Bcl2a1b*, *Trem2*, *Cd68*, *Ctsd*, *C1qa*) show stronger enrichment near plaques in clustered environments. These are the immune/DAM genes whose proximity signal is amplified by surrounding plaques.
- Genes **below the diagonal** (e.g., *App*, *Snap25*, *Grin2a*, *Slc17a7*, *Bin1*) show stronger enrichment near plaques in isolated environments. These tend to be neuronal genes whose spatial relationship to plaques may be confounded by tissue architecture in dense plaque regions.

This divergence provides direct evidence that the transcriptional response to plaque proximity is qualitatively different depending on the local plaque density: isolated plaques elicit a neuron-dominated spatial signature, while clustered plaques activate a stronger immune/microglial programme.

# Summary

```{r}
#| label: summary-table
summary_df <- tibble(
  metric = c(
    "Total cells",
    "Isolated cells (≤1 plaque within 100 µm)",
    "Clustered cells (≥2 plaques within 100 µm)",
    "Genes tested in interaction model",
    "Genes with significant interaction (BH < 0.05)",
    "Median interaction coefficient (significant genes)",
    "Genes with consistent sign across all 6 brains"
  ),
  value = c(
    nrow(cell_distances_all),
    sum(cell_distances_all$density_class == "isolated"),
    sum(cell_distances_all$density_class == "clustered"),
    nrow(interaction_results),
    sum(interaction_results$sig),
    interaction_results |> filter(sig) |> pull(int_estimate) |> median() |> signif(3),
    per_brain_results |>
      group_by(gene) |>
      summarise(all_same_sign = n_distinct(sign(estimate)) == 1, .groups = "drop") |>
      filter(all_same_sign) |>
      nrow()
  )
)

summary_df |>
  kbl(caption = "Multi-plaque neighbourhood analysis summary") |>
  kable_styling("striped", full_width = FALSE)
```

**Key takeaways:**

1. **Plaque density is spatially heterogeneous.** Two-thirds of cells (~242k) reside in clustered plaque environments (>=2 plaques within 100 µm), with pronounced hotspots visible in the spatial maps. This heterogeneity means that the "distance to nearest plaque" metric used in earlier analyses conflates proximity with density.

2. **DAM gene expression scales with plaque density.** All six DAM markers tested (*Cst7*, *Trem2*, *Cd68*, *Il1b*, *Lyz2*, *Bcl2a1b*) show higher expression in clustered vs isolated environments, consistent with a cumulative amyloid burden hypothesis where multiple nearby plaques elicit a stronger microglial activation programme.

3. **Distance x density interactions are pervasive and biologically consistent.** 437 of 459 genes (95%) show a significant interaction, stable across density radii (50--150 µm) and consistent in sign across all 6 brains. Positive interactions (e.g., *App*, *Bace1*, *Bin1*) indicate amplified proximity effects in dense regions; negative interactions (e.g., *C1qa*, *Bcl2a1b*, *Aif1*) suggest saturation of the immune response in heavily burdened tissue.

4. **Isolated and clustered environments activate distinct spatial gene programmes.** Near isolated plaques, the dominant spatial signal is neuronal (*Snap25*, *Slc17a7*, *Grin2a*); near clustered plaques, canonical DAM/immune genes (*Cst7*, *Trem2*, *Lyz2*) dominate. This qualitative shift suggests that the microglial DAM response requires a threshold level of local amyloid burden to fully engage.
